### 2024-12-04 paper/paper.qmd
final commit

---
title: "Mountaineering: Be lucky first, prepared second"
subtitle: "Preperation and Timing are more important than Gender Identity, Age, Amount of expedition members, and height of peak when mountaineering in the Himalayas."
author: 
  - Daniel Xu
thanks: "Code and data are available at: [https://github.com/possibleburger2/](https://github.com/possibleburger2/Himalayan-Climbing-Expeditions-Data-Analysis)."
date: today
date-format: long
abstract: "This study asseses factors that contribute to success in Himalayan Expeditions from Himalayan Expedition data from 1905-2019"
format: pdf
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(readr)
library(dplyr)
library(arrow)
library(here)
library(ggplot2)
library(arkhe)
library(modelsummary)
library(kableExtra)

# Load data
peaks <- read_parquet(here("data/02-analysis_data/peaks_analysis_data.parquet"))
expeditions <- read_parquet(here("data/02-analysis_data/expeditions_analysis_data.parquet"))
members <- read_parquet(here("data/02-analysis_data/members_analysis_data.parquet"))
```

# Introduction

Mountaineering has been growing as an activity since the 90s, the views and sense of challenge have brought people around the globe to summit more and more mountains every year. It is no surprise then, that when combined with a mountain range containing the highest above sea level peak in the world the Himalayas have had a rich documented history of mountaineering attempts. Mountaineering is a dangerous hobby however, with unpredictable weather and temperatures so high up it is natural that not all attempts to summit mountain peaks succeed here. This paper will find what factors have affected the climbers of the past to either conquer or be conquered.

This paper uses data sourced from @Cookson, based on The Himalayan Database [@Salisbury] about expeditions from 1905-2019 to analyze which factors help or deter climbers from succeeding. We use a binomial logistic regression  model to determine the probability of success based on a variety of demographic, enviromental, and organizational factors.

The estimand is the relationship between said factors and the success of a mountaineering expedition. By analyzing the data set, we can identify which factors contribute to a success or failure and by how much.

My analysis shows that Mountaineering is positively impacted by oxygen used, younger age, going in the fall and spring, and being a man while is is negatively impacted by the height of the mountain and the amount of people not hired in the expedition.

These findings practically inform climbers about the impact of good preparation and timing over personal factors like age and sex, showing that regardless of who you are, what you do is the biggest factor in succeeding in high-altitude mountaineering.

The paper is further organized into four sections. @sec-data discusses how the dataset used for the analysis was obtained and pre-processed. I will explain the variables of interest in the dataset used for the analysis. @sec-model describes the model being used for the analysis. @sec-results then highlights and discusses the trends and associations found during the analysis. Lastly, @sec-discussion talks about some interesting trends found in @sec-results in depth, linking it to the real world and also highlights the weaknesses and future of my analysis.

# Data {#sec-data}

The datasets were cleaned and analysed using the statistical programming software `R` [@citeR] along with the help of `tidyverse` [@citeTidyverse], `knitr` [@citeKnitr], `ggplot2` [@citeGgplot], `here` [@citeHere], `dplyr` [@citeDplyr], `rstanarm` [@citeRstanarm], `arrow` [@citeArrow] and `kableExtra` [@citeKableExtra].

## Overview
The raw datasets were obtained from @Cookson, who sourced them from The Himalayan Database [@Salisbury].

The Himalayan Database is a compilation of records for all expeditions that have climbed in the Nepal Himalaya. The database is based on the expedition archives of Elizabeth Hawley, a long-time journalist based in Kathmandu, and it is supplemented by information gathered from books, alpine journals and correspondence with Himalayan climbers.

The original database currently covers all expeditions from 1905 through Spring-Summer 2023 to the most significant mountaineering peaks in Nepal. Also included are expeditions to both sides of border peaks such as Everest, Cho Oyu, Makalu and Kangchenjunga as well as to some smaller border peaks. Data on expeditions to trekking peaks are included for early attempts, first ascents and major accidents. The updates to this database are published bi-annually.

My dataset, derived from Cookson's, contains the entries from 1905 through Spring 2019.

A person becomes an entry in my analysis dataset if, between 1905 and Spring 2019, they attempted to climb any one of the many Himalayan peaks in Nepal.

The Peaks data includes several variables. The "peak_id" is a unique identifier for each peak. "peak_name" refers to the common name of the peak, such as "Mount Everest." "peak_alternative_name" is the alternative name of the peak, like "Sagarmatha" for Mount Everest in Nepal. "height_metres" is the height of the peak in metres, and "climbing_status" indicates whether the peak has been climbed or not. The "first_ascent_year" is the year when the peak was first successfully climbed, and "first_ascent_country" shows the country (or countries) of the climbers who made the first ascent. The "first_ascent_expedition_id" is a unique identifier for the expedition that achieved the first ascent.

The Expeditions data contains variables such as "expedition_id," which is a unique identifier for each expedition. "peak_id" links to the unique identifier for the peak being climbed, and "peak_name" refers to the common name of the peak. "year" is the year the expedition took place, and "season" indicates the season when the expedition occurred (such as Spring or Summer). The "basecamp_date" is the date the expedition arrived at basecamp, and "highpoint_date" is the date the expedition reached the highest point or summit of the peak. "termination_date" shows when the expedition ended, and "termination_reason" describes why the expedition was called off. "highpoint_metres" is the highest elevation reached by the expedition, while "members" refers to the number of people in the expedition, typically the number of foreigners in Nepal or non-hired members in China. "member_deaths" is the number of members who died, and "hired_staff" is the number of hired staff who went above basecamp. "hired_staff_deaths" shows the number of hired staff who died, and "oxygen_used" indicates whether any member of the expedition used oxygen. "trekking_agency" is the name of the agency organizing the expedition.

The Members data includes variables such as "expedition_id," which is a unique identifier for the expedition, linking to the Expeditions data. "member_id" is a unique identifier for each person, though it may differ across expeditions. "peak_id" links to the unique identifier for the peak being climbed, and "peak_name" is the common name of the peak. "year" is the year of the expedition, and "season" refers to the season of the expedition. "sex" is the sex of the person, and "age" is the person's age during the expedition based on available data (such as summit date, date of death, or basecamp arrival). "citizenship" is the person's nationality, and "expedition_role" is their role in the expedition. "hired" indicates whether the person was hired by the expedition, and "highpoint_metres" is the highest point reached by the person. "success" shows whether the person succeeded in summiting the main peak or a sub-peak, depending on the expedition's goal. "solo" indicates whether the person attempted a solo ascent, and "oxygen_used" shows whether the person used oxygen. "died" indicates whether the person died during the expedition, while "death_cause" shows the primary cause of death. "death_height_metres" is the height at which the person died, and "injured" indicates whether the person was injured. "injury_type" refers to the main cause of injury, and "injury_height_metres" is the height at which the injury occurred.

```{r, fig.width = 12, fig.height = 6}
#| warning: false
#| message: false
#| echo: false
#| label: fig-ssdistribution
#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."

peaks |>
  ggplot(aes(height_metres)) +
  geom_histogram(
    binwidth = 200,
    alpha = 0.8
  ) +
  annotate("text", 8450, 17, label = "Mount Everest") +
  annotate(
    "curve",
    x = 8500,
    y = 15,
    xend = 8775,
    yend = 2,
    curvature = -0.25,
    arrow = arrow(length = unit(2, "mm"))
  ) +
  labs(
    title = "Distribution of Himalayan Peak Heights",
    caption = "Source: The Himalayan Database",
    x = "Height (m) above sea level",
    y = "Number of peaks"
  )
```

@fig-ssdistribution shows most peaks sit around 6000 meters above sea level, with a right skewed distribution and a maximum of almost 9000 meters and a minimum of 5250 meters.

The Himalayan mountain range spans X KM (source), which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas. Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig-cap: "Amount of unclimbed Himalayan peaks"

peaks |>
  ggplot(aes(climbing_status, fill = climbing_status)) +
  geom_bar() +
  labs(
    title = "More than a quarter of Himalayan peaks remain unclimbed",
    x = "",
    y = "Number of peaks"
  ) +
  theme(legend.position = "none")
```
This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.

\newpage
Why do so many climbs fail however? The following graph shows the most common reasons.

```{r,fig.width = 12, fig.height = 9}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-termination
#| tbl-cap: "Top 5 most frequent pollsters, with count of polls, average pollscore (lower scores indicate less bias), and average numeric grade (higher values indicate greater reliability)."

filtered_expeditions <- expeditions %>%
  filter(!termination_reason %in% c("Success (main peak)", "Success (subpeak)", "Success (claimed)"))
ggplot(filtered_expeditions, aes(x = termination_reason)) +
  geom_bar(fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Count of Expeditions by Termination Reason",
    x = "Termination Reason",
    y = "Count"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches. 
\newpage

```{r, fig.width = 12, fig.height = 3}
#| warning: false
#| message: false
#| echo: false
#| label: fig-summit
#| fig-cap: "Distribution of First ascents across years"

peaks |>
  ggplot(aes(first_ascent_year)) +
  geom_histogram(
    binwidth = 5,
    alpha = 0.8
  ) +
  scale_x_continuous(breaks = seq(1910, 2020, 10)) +
  labs(title = "Climbers are still summitting peaks for the first time",
       subtitle = "Year of first ascen",
       x = "Year of first ascent",
       y = "Number of first ascents")
        panel.grid.minor = element_blank()
```

@fig-summit shows that while there was an intial surge then drop off of first ascents, there was again a steady increase in first ascents actually peaking in 2019, and the 2020 data was inconclusive since not all of the year was recorded in the data. This shows that first ascents are currently trending upwards
\newpage

```{r,fig.width = 12, fig.height = 6}
#| warning: false
#| message: false
#| echo: false
#| label: fig-trend
#| fig-cap: "Distribution of first ascents by country of citizenship"

top_20_countries <- peaks |>
  filter(!is.na(first_ascent_country)) |>
  separate_rows(first_ascent_country, sep = ",") |>
  mutate(
    # Get rid of whitespace after separate_rows
    first_ascent_country = str_squish(first_ascent_country),
    # Aggregate W Germany and Germany into "Germany"
    first_ascent_country = ifelse(
      first_ascent_country == "W Germany",
      "Germany",
      first_ascent_country
    )
  ) |>
  dplyr::count(first_ascent_country, name = "first_ascents", sort = TRUE) |>
  mutate(first_ascent_country = fct_reorder(first_ascent_country, first_ascents)) |>
  dplyr::slice_max(order_by = first_ascents,n=20)

top_20_countries |>
  ggplot(aes(first_ascents, first_ascent_country, fill = first_ascent_country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 150, 25)) +
  labs(
    title = "Nepal and Japan lead the way in first ascents",
    subtitle = "First ascents a country's citizen was involved in",
    caption = "Source: The Himalayan Database",
    x = "Number of first ascents",
    y = ""
  ) 
```

@fig-trend Expectly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..
\newpage

```{r}
#| echo: false
#| warning: false
#| message: false
#| inlcude: false
#| label: fig-country
countries_by_decade <- peaks |>
  dplyr::filter(!is.na(first_ascent_country), first_ascent_year >= 1910) |>
  tidyr::separate_rows(first_ascent_country, sep = ",") |>
  dplyr::mutate(
    first_ascent_country = stringr::str_squish(first_ascent_country),
    first_ascent_country = ifelse(
      first_ascent_country == "W Germany",
      "Germany",
      first_ascent_country
    ),
    first_ascent_decade = first_ascent_year %/% 10 * 10,
    first_ascent_country = forcats::fct_lump(first_ascent_country, 8)
  ) |>
  dplyr::count(first_ascent_country, first_ascent_decade, name = "first_ascents") |>
  dplyr::group_by(first_ascent_decade) |>
  dplyr::mutate(pct_of_ascents = first_ascents / sum(first_ascents)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    first_ascent_country = forcats::fct_reorder(first_ascent_country, -first_ascents, sum),
    first_ascent_country = forcats::fct_relevel(first_ascent_country, "Other", after = Inf)
  )

countries_by_decade |>
  ggplot(aes(first_ascent_decade, pct_of_ascents, fill = first_ascent_country)) +
  geom_col() +
  scale_x_continuous(breaks = seq(1930, 2010, 20)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +  # Correct label function
  facet_wrap(~ first_ascent_country) +
  labs(
    title = "Nepal has been consistently involved in first ascents",
    subtitle = "Percent of first ascents involving a countries' citizens",
    x = "Decade of first ascent",
    y = ""
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(colour = "black"),
    strip.background = element_blank()
  )
```
Notably there was only one data point past 1930, where X expedition had an ascent for X mountain in 1905 that was left out to save space from the UK so there should be a single bar only for the UK at 100% at 1905.

## Measurement

## Model {#sec-model}
I used a Binary Logistic Regression uses the age of an individual,the amount of members in their expedition, the amount of oxygen they used, the season of their climb, their sex and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding". A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success"
The logistic regression model I will be using is:
$$
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} + \beta_5 \times \text{sex} + \beta_6 \times \text{heightmetres} 
$$

## Model Justification

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary1

# Load your models
model1 <- readRDS(here("models/binary_logit.rds"))
model2 <- readRDS(here("models/original_model.rds"))

modelsummary(model2,statistic = c("p.value"))
```
\newpage
\newpage

shows the influence (top number) and P-Value (bottom number in brackets) of each variable. For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship. A P-value of 0.001 means only 0.1% of the time this relationship can be explained by random chance. Because the standard level of uncertainty is less than 5% we should try to remove variables over a 0.05 P-value. There are also several figures relating to model accurary at the bottom of the figure;the main one to focus on is RMSE. An RMSE of 0.41 means the model predicts results within 41% of the actual results. For a model with a binary outcome it means our model predicts successes 59% of the time.]

Because we have so many statistically insignificant variables at a 5% uncertainty we should try to remove them to see the impact it has on the model
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary2

modelsummary(model1, statistic = c("p.value"))
```
\newpage
\newpage

shows us that despite removing citizenship and hired_staff the RMSE only grains 0.01, meaing it loses 1% accuracy. This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
```{r, fig.width = 2, fig.height = 1}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

results <- anova(model1,model2,
      test = "LRT")
kable(results, caption = "ANOVA Model Comparison") |>
  kable_styling(full_width = FALSE, position = "center")
```
This anova table just confirms that there is no difference in the modelling ability after removing the two predictor variables.

## Results {#sec-results}
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-model-coefficients
#| fig-cap: Coefficients of the model

# Coefficients for the model
coefficients <- broom::tidy(model1, conf.int = TRUE)

# Plot for the coefficients of the model
ggplot(coefficients, aes(x = estimate, y = term)) +
  geom_point() + # Coefficient points
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) + # Confidence/Credible interval bars
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") + # Add zero line
  labs(x = "Coefficient Estimate", y = "Predictor Variable") + # Axis labels
  theme_minimal() # Minimal theme for better readability
``` 
@fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success. The season you go are all relative to fall, so going during the fall has a large measured impact on success as well as oxygen used and sex.

# Discussion {#sec-discussion}

## Climbing success relys on oxygen used signifigantly

There is a large certain positive affect of oxygen used on a climb, this can be due to many reasons. While it is definitely true that oxygen will help with a large obstacle called altitude sickness it also represents a change in technology and climbing culture. As technology has gotten better it has been easier and cheaper to carry more oxygen, also due to an ever increasing use of sherpas or hired help to reach summits. These are all issues that also increase summit success rates so the data may be over representing the actual effectiveness of oxygen. What it also represents is the overall preparedness of the climber, and that will naturally influence the success rate.

## Spring and Autumn are the best time for expeditions

Autumn and spring, as transitional seasons, provide favorable conditions for mountaineering expeditions. The periods offer moderate temperatures and more stable weather patterns compared to the extremes of summer and winter. @fig-model-coefficients  show the difference in probabilities of success by seasons and highlight higher chances of success in spring and autumn. @Season provides tips for summit success when climbing Mt. Everest and suggest that the best time to climb it is in the Autumn/ Fall season. This is in-line with our findings.

Conversely, summer and winter present significant challenges for mountaineers. Summer brings the risks of intense heat and thunderstorms, increasing the likelihood of rockfall and other hazards while winter conditions are characterized by extreme cold, high winds, heavy snowfall, and increased avalanche danger, making ascent and descent more difficult and hazardous. As a result, winter is generally considered the least desirable season for mountaineering expeditions.

In conclusion, autumn and spring are the optimal seasons for mountaineering expeditions, and season selection is an important factor in expedition planning.

## Chances of success decline with increase in age
Age is a significant factor in mountaineering success, with success rates generally decreasing as climbers get older. This decline can be attributed to various age-related factors, such as the natural loss of muscle mass, strength, and aerobic capacity. As climbers age, they may also experience reduced agility, flexibility, and slower recovery, which can hinder their performance on difficult terrain and in recovering from intense physical exertion.

In addition to these physical challenges, older climbers may face health issues like cardiovascular problems, joint stiffness, and decreased bone density, all of which can further limit their ability to meet the physical demands of mountaineering. These factors contribute to a reduced ability to handle the stresses of expeditions and increase the likelihood of difficulties along the way.

However, it is important to recognize that success in mountaineering is not solely determined by physical ability. Experience, skill, preparation, and mental strength play crucial roles in overcoming the challenges of the climb. Older climbers, despite the physiological decline that comes with age, can still achieve success by leveraging their experience, using strategic planning, and maintaining a strong mindset. These factors enable them to adapt to the challenges they face, showing that age does not necessarily preclude achieving mountaineering goals.

In summary, age-related impacts on mountaineering success are significant, with success rates generally decreasing as climbers age. However, older climbers can overcome some of these challenges through experience, skill, strength training, and a strong mindset, allowing them to continue enjoying mountaineering well into their later years.

## Weaknesses and Limitations
While my analysis offers valuable insights into the factors influencing a successful summit, it is important to acknowledge the weaknesses and limitations.

In summary, while age can impact mountaineering success, older climbers can still achieve summit success through a combination of experience, skill, and mental strength.

The analysis conducted has some limitations that need to be addressed. One of the main weaknesses is that the success probabilities may be skewed due to inaccuracies in the data. For example, the number of solo ascents in the dataset is much smaller than the number of group ascents, yet the model suggests that solo ascents have higher success rates. This could lead to a misconception that solo climbs are more successful, without considering other contributing factors such as the determination and willpower of those opting for solo expeditions. These climbers might be more motivated, which could be a confounding factor influencing their success rates. The dataset also has some inconsistencies, especially in variables like nationality and group size. Inconsistent reporting methods between different countries, such as the way group size is recorded in Nepal versus China, led to these factors being removed from the analysis. This inconsistency limits the ability to draw conclusions based on these variables.

Additionally some variables were not readily available in the data set that would have a large influence on the findings. Factors like individual skill, mental direction, and external factors like difficulty of the climb are not included or well represented in the data.
## Future Directions
Looking ahead, future research should focus on improving the dataset by exploring additional factors that could influence summit success. One such factor is the number of attempts made by a climber. Understanding whether repeated attempts increase the likelihood of success or if there is a point of diminishing returns could provide valuable insights into the role of perseverance and learning from past experiences. Additionally, examining the level of technical skill or experience of climbers could help determine if those with more specialized training or knowledge are more likely to succeed. A deeper understanding of the climbers' preparedness—both physically and mentally—would also be crucial. This could include factors like training regimes, mental resilience, and coping strategies, all of which may play a critical role in a climber's ability to reach the summit. Lastly, studying the support systems available to climbers, such as the role of guides, sherpas, or expedition teams, could reveal how these resources impact success rates. By considering these factors, future research can provide a more comprehensive understanding of what contributes to successful mountaineering expeditions.

\newpage

\appendix

# Appendix {#sec-appendix}
## Cleaning
For the analysis data, the cleaning steps I took were:

The data cleaning process described in the provided script involves transforming raw data into a more consistent, accurate, and analysis-ready format. The first step in the cleaning process is applied to the peaks dataset, which contains information about various mountain peaks. In this step, columns are renamed for clarity, such as changing PEAKID to peakid and PKNAME to peakname. Only the necessary columns are retained, while others are discarded to simplify the dataset. A key aspect of this cleaning is handling the climbingstatus column. The raw data uses numerical values (0, 1, 2) to represent the climbing status, so these numbers are recoded into meaningful labels like "Unknown," "Unclimbed," and "Climbed." Additionally, a data entry error is corrected for Sharpu II (identified by peakid == "SPH2"), where the year of the first ascent is changed to 2018 to fix an obvious mistake.

The second dataset cleaned is the expeditions dataset, which tracks various climbing expeditions. Here, the dataset is joined with the peaks dataset to bring in the peakname for each expedition using the peakid. Only the relevant columns are selected and renamed for clarity. The highpointmetres column is adjusted by setting any value of 0 to NA, as it likely indicates missing data. Categorical variables like terminationreason and season are recoded to make them more readable. The terminationreason column is recoded from numeric values (such as 0, 1, and 2) into descriptive labels like "Success (main peak)," "Bad weather," and "Accident." Similarly, the season column is recoded from numbers (0-4) into readable season names like "Spring," "Summer," "Autumn," and "Winter."

The third dataset, members, contains information about individual climbers. This dataset is also joined with the peaks dataset to associate each climber's expedition with the corresponding peak name. As with the other datasets, only the relevant columns are retained and renamed. Any value of 0 in the age column is treated as missing data and changed to NA. The deathcause and injurytype columns, which are initially represented by numeric values (0-12), are recoded into readable categories like "AMS" (Acute Mountain Sickness) and "Exhaustion." The cleaning also involves handling information related to deaths and injuries. For climbers who have died, the deathcause and deathheightmetres are updated, and similarly, the injurytype and injuryheightmetres are updated for injured climbers.

# References

### 2024-12-04 scripts/00-simulate_data.R

@@ -17,14 +17,18 @@ set.seed(853)
 n_peaks <- 20
 n_members <- 100
 n_expeditions <- 10
-# Simulation parameters 
+# Simulation parameters
+
 Average_height <- 6100  # height in meters
-Tallest_peak <- 8849 # We can easily include the highest peak, since we know the tallest part of the tallest mountain,
- #but since the lowest peak of a mountain must be 1000 feet up from its base, it is difficult to determine the lowest peak in the Himalayas
+Tallest_peak <- 8849 # We can easily include the highest peak, since we know 
+#the tallest part of the tallest mountain,
+#but since the lowest peak of a mountain must be 1000 feet up from its base, 
+#it is difficult to determine the lowest peak in the Himalayas
 
 #create Weighted list of Seasons
-seasons <- c("Spring","Summer", "Autumn", "Winter")#weighted since people more often do ascents during warmer seasons/seasons with calmer weather
-seasons_prob <- c(0.35,0.2,0.35,0.1)
+seasons <- c("Spring", "Summer", "Autumn", "Winter")#weighted since people more 
+#often do ascents during warmer seasons/seasons with calmer weather
+seasons_prob <- c(0.35, 0.2, 0.35, 0.1)
 
 #create List of Peaks along with corresponding peak id and height
 peaks <- data.frame(
@@ -36,11 +40,15 @@ peaks <- data.frame(
     "Annapurna II", "Annapurna III", "Annapurna IV", "Langtang Lirung", 
     "Ganesh I", "Pumori"
   ),
-  peak_alt = c("Sagarmatha","Chhogori",NA,"E1","Kamalung",NA,NA,NA,NA,"Hidden Peak",NA,NA,"Gosainthan","Diamir",NA,NA,NA,NA,NA,"Everest's Daughter"),
+  peak_alt = c("Sagarmatha", "Chhogori", NA, "E1", "Kamalung", NA, NA , NA, NA,
+               "Hidden Peak", NA, NA, "Gosainthan", "Diamir", NA, NA, NA, NA, 
+               NA, "Everest's Daughter"),
   peak_height = (5000 + 4000 * rbeta(n_peaks,2,5)),
-  climbing_status = sample(c("Climbed", "Unclimbed"),n_peaks,replace = TRUE, prob = c(0.8,0.2)),
+  climbing_status = sample(c("Climbed", "Unclimbed"), n_peaks, 
+                           replace = TRUE, prob = c(0.8,0.2)),
[…]

### 2024-12-04 scripts/03-clean_data.R

@@ -161,12 +161,3 @@ members <-
 write_parquet(expeditions,"data/02-analysis_data/expeditions_analysis_data.parquet")
 write_parquet(members,"data/02-analysis_data/members_analysis_data.parquet")
 write_parquet(peaks,"data/02-analysis_data/peaks_analysis_data.parquet")
-
-write_csv(expeditions,"data/02-analysis_data/expeditions_analysis_data.csv")
-write_csv(members,"data/02-analysis_data/members_analysis_data.csv")
-write_csv(peaks,"data/02-analysis_data/peaks_analysis_data.csv")
-members_exp <- members %>%
-  left_join(expeditions %>%
-              select(expedition_id, basecamp_date, termination_date, members, hired_staff),
-            by = "year")
-view(members_exp)

### 2024-12-04 scripts/06-model_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
-# Purpose: Models winner of 2024 US election
-# Author: Daniel, Vandan
+# Purpose: Models climbing successes
+# Author: Daniel
 # Date: 4 November 2024
 # Contact: danie.xu@mail.utoronto.ca
 # License: MIT
@@ -18,10 +18,7 @@ library(stats)
 set.seed(42)  
 
 #### Read the data and create model ####
-# Read the cleaned analysis dataset
-#expeditions <- read_csv("data/02-analysis_data/expeditions_analysis_data.csv")
-#members <- read_csv("data/02-analysis_data/members_analysis_data.csv")
-#peaks <-read_csv("data/02-analysis_data/peaks_analysis_data.csv")
+
 expeditions <- read_parquet("data/02-analysis_data/expeditions_analysis_data.parquet")
 members <- read_parquet("data/02-analysis_data/members_analysis_data.parquet")
 peaks <- read_parquet("data/02-analysis_data/peaks_analysis_data.parquet")
@@ -48,9 +45,10 @@ model <- glm(
   formula = success ~ age + members + hired_staff + oxygen_used.x + season + height_metres + citizenship + sex,
   data = test_data,
   family = "binomial")
+
 summary(model)
 
-# model with hired_staff and citenzenship removed 
+# model with hired_staff and citizenship removed 
 model2 <- glm(formula = success ~ age + members + oxygen_used.x + season +height_metres + sex,
               data = test_data,
               family = "binomial")
@@ -67,4 +65,8 @@ saveRDS(
   file = "models/binary_logit.rds"
 )
 
+saveRDS(
+  model,
+  file= "models/original_model.rds"
+)
 

### 2024-12-04 paper/paper.qmd
Save paper, code style changes to model data

@@
 
-The Himalayan mountain range spans X KM (source), which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas.
-Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.
+The Himalayan mountain range spans a large amount of distance, which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas.
+Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above 2000m sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.
+
+\newpage
 
@@
 This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
-
+Why do so many climbs fail however? The following graph shows the most common reasons.
 \newpage
-Why do so many climbs fail however? The following graph shows the most common reasons.
 
@@
 #| label: tbl-termination
-#| tbl-cap: "Top 5 most frequent pollsters, with count of polls, average pollscore (lower scores indicate less bias), and average numeric grade (higher values indicate greater reliability)."
+#| tbl-cap: "Common reasons why attempts fail"
 
@@
 \newpage
-
 ```{r, fig.width = 12, fig.height = 3}
@@
 This shows that first ascents are currently trending upwards
+
 \newpage
 
-```{r,fig.width = 12, fig.height = 6}
+```{r,fig.width = 12, fig.height = 5}
 #| warning: false
@@
 
-@fig-trend Expectly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..
+@fig-trend Expectedly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..
+
 \newpage
@@
 
+\newpage
+
 ## Measurement
+The measurement of Himalayan climbing data in the referenced dataset involves converting real-world phenomena—such as expeditions, summits, and fatalities—into structured entries.
+This process begins with records collected from expedition reports, which are typically submitted by climbers or organizations.
+These reports include details about expedition dates, routes, climbers' nationalities, success in reaching the summit, and any fatalities or injuries that occurred.
+
+The dataset relies heavily on sources like The Himalayan Database, a comprehensive archive maintained by Elizabeth Hawley and her successors.
+This database aggregates information from direct interviews with climbers, expedition documentation, and third-party reports.
+However, the translation of these real-world events into dataset entries introduces several challenges.
+Discrepancies may arise due to incomplete or inconsistent reporting, varying definitions of a “summit,” and the underrepresentation of smaller, independent expeditions.
+
+For example, an expedition is recorded in the dataset only if sufficient documentation is available to verify its occurrence and outcomes.
+Factors such as weather conditions, local incidents, or logistical difficulties that don't result in a summit or a fatality may not be consistently recorded.
+Additionally, subjective elements, like the climbers’ or organizers' interpretation of success, influence the data's structure.
+These steps, while thorough, highlight potential biases in how phenomena are represented as measurable data.
 
@@
 ## Model Justification
-
 ```{r}
@@
 
-modelsummary(model2,statistic = c("p.value"))
+#modelsummary(model2,statistic = c("p.value")) simply took output and inputted as png, the length of the table is too long and won't wrap
 ```
 \newpage
-\newpage
 
-shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
+![](image1.png)
+This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
 For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship.
@@
 An RMSE of 0.41 means the model predicts results within 41% of the actual results.
-For a model with a binary outcome it means our model predicts successes 59% of the time.]
+For a model with a binary outcome it means our model predicts successes 59% of the time.(removed entries for each country due to formatting issues)
 
@@
 \newpage
-\newpage
-
-shows us that despite removing citizenship and hired_staff the RMSE only grains 0.01, meaing it loses 1% accuracy.
+The second table shows the results when citizenship and hired_staff are removed.
+We see the RMSE only grains 0.01, meaning it loses 1% accuracy.
 This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
@@
 ## Cleaning
-For the analysis data, the cleaning steps I took were:
+For the analysis data, the cleaning steps were from @Cookson
 
@@
 
+## Observational data
+Himalayan climbing data, particularly concerning summits and fatalities on peaks like Mount Everest, may misrepresent actual outcomes due to several factors, including incomplete reporting, commercial biases, and evolving definitions of success.
+These discrepancies can create a skewed understanding of the risks and achievements in high-altitude mountaineering.
+Incomplete Reporting
+
+Many summits, fatalities, and near-misses in the Himalayas go unreported or inaccurately documented due to the remote and challenging environments of the region.
+Smaller expeditions, particularly those from less prominent countries or by independent climbers, may not submit detailed records.
+Additionally, incidents on less-frequented peaks are less likely to be covered in international or even regional climbing databases, leading to underrepresentation in global statistics.
+For example, organizations like the Himalayan Database, while comprehensive, rely heavily on self-reported data, which can vary in reliability depending on the source.
+Commercial Influences
+
+The rise of commercial expedition companies has significantly influenced how climbing data is presented.
+For marketing purposes, these companies may highlight success rates while downplaying or omitting fatalities or unsuccessful attempts.
+Moreover, certain deaths that occur before or after the official climbing season might not be included in annual tallies, further skewing statistics.
+High-profile peaks like Everest often see overrepresentation in the data, as their fame and accessibility make them more frequently reported than other Himalayan mountains.
+
+The definition of a “summit” itself can vary, adding further complications.
+On Everest, for example, some climbers might claim summits upon reaching the South Summit or other points close to the highest peak, particularly in cases of bad weather or emergencies.
+Such claims are sometimes accepted without verification, leading to inflated success statistics.
+Disputes over what constitutes a legitimate ascent, particularly as aided by modern technology like oxygen tanks and fixed ropes, also muddy the waters of accurate representation.
+Bias in Reporting Fatalities
+
+Fatality rates may also be misrepresented due to a focus on high-profile accidents involving Western climbers or commercial expeditions, while incidents involving Sherpas and local climbers are often underreported.
+Sherpas, who take on the riskiest roles during climbs, such as route preparation and heavy load carrying, face disproportionate risks that may not be fully reflected in the data.
+Conclusion
+
+The misrepresentation of Himalayan climbing data has implications for understanding the true risks and challenges of high-altitude mountaineering.
+Efforts to improve the accuracy of this data, such as independent verification and more comprehensive reporting systems, are essential for providing climbers, researchers, and policymakers with reliable information.
+Sources like The Himalayan Database [@Salisbury], the American Alpine Journal, and recent academic studies on mountaineering ethics and statistics provide valuable insights but also highlight the need for more transparent and standardized data collection practices.
+
 # References

### 2024-12-04 scripts/06-model_data.R

@@ -25,7 +25,8 @@ peaks <- read_parquet("data/02-analysis_data/peaks_analysis_data.parquet")
 
 #merge data
 
-members_exp <-  merge(members,expeditions,by = c("expedition_id","year","season","peak_id","peak_name"),all.x = TRUE)
+members_exp <-  merge(members,expeditions,by = c("expedition_id","year","season",
+                                            "peak_id","peak_name"),all.x = TRUE)
 total <- merge(members_exp,peaks, by = c("peak_id","peak_name"))
 
 #split data for testing
@@ -42,14 +43,16 @@ nrow(test_data)   # Testing data size
 
 # Fit a Bayesian logistic regression model
 model <- glm(
-  formula = success ~ age + members + hired_staff + oxygen_used.x + season + height_metres + citizenship + sex,
+  formula = success ~ age + members + hired_staff + oxygen_used.x + season 
+  + height_metres + citizenship + sex,
   data = test_data,
   family = "binomial")
 
 summary(model)
 
 # model with hired_staff and citizenship removed 
-model2 <- glm(formula = success ~ age + members + oxygen_used.x + season +height_metres + sex,
+model2 <- glm(formula = success ~ age + members + oxygen_used.x + season 
+              + height_metres + sex,
               data = test_data,
               family = "binomial")
 

### 2024-12-04 scripts/00-simulate_data.R
Update 00-simulate_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Simulates polling data
+# Purpose: Simulates climbing data
 # Author: Daniel Xu
 # Date: 4 November 2024
 # Contact: danie.xu@mail.utoronto..ca

### 2024-12-04 paper/paper.qmd
update paper

@@
 date-format: long
-abstract: "This study asseses factors that contribute to success in Himalayan Expeditions from Himalayan Expedition data from 1905-2019"
+abstract: "This study analyzes over a century of Himalayan mountaineering expeditions (1905–2019) to identify key factors influencing success.
+Using logistic regression, we find that preparation and timing—such as the use of supplemental oxygen and climbing during optimal seasons—are more important predictors of success than individual traits like age, gender, or the height of the peak.
+These findings reveal that success in high-altitude climbing depends more on strategic choices than personal attributes.
+This insight underscores the value of informed planning and environmental awareness for achieving mountaineering goals."
 format: pdf
@@
 
-## Climbing success relys on oxygen used signifigantly
+## Climbing success relys on oxygen used significantly
 

### 2024-12-14 paper/paper.qmd
update clean data

@@
 bibliography: references.bib
-
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
 ## Overview
+
 The raw datasets were obtained from @Cookson, who sourced them from The Himalayan Database [@Salisbury].
@@
 "injury_type" refers to the main cause of injury, and "injury_height_metres" is the height at which the injury occurred.
+
+Members of expedition
+
+```{r}
+
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-members
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+expeditions |>
+  ggplot(aes(members)) +
+  geom_histogram(
+    binwidth = 2,
+    alpha = 0.8
+  ) +
+  labs(
+    title = "Distribution of Expedition members",
+    caption = "Source: The Himalayan Database",
+    x = "Height (m) above sea level",
+    y = "Number of peaks"
+  )
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-members
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+expeditions$success <- expeditions$termination_reason %in% c("Success (main peak)", "Success (subpeak)")
+success_rate <- expeditions |>
+  group_by(members)|>
+  summarize(success_rate = mean(success))
+ggplot(success_rate, aes(x = members, y = success_rate)) +
+  geom_line(color = "blue") + # Line plot
+  geom_point(color = "red") + # Points to emphasize values
+  labs(
+    title = "Success Rate by Members",
+    x = "Number of Members",
+    y = "Success Rate"
+  ) +
+  theme_minimal()
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-season
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions,aes(season)) +
+  geom_bar()
+```
 
@@
 ```
+
 This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
-Why do so many climbs fail however? The following graph shows the most common reasons.
+Why do so many climbs fail however?
+The following graph shows the most common reasons.
 \newpage
@@
 \newpage
+
 ```{r, fig.width = 12, fig.height = 3}
@@
 ```
+
 Notably there was only one data point past 1930, where X expedition had an ascent for X mountain in 1905 that was left out to save space from the UK so there should be a single bar only for the UK at 100% at 1905.
@@
 ## Measurement
+
 The measurement of Himalayan climbing data in the referenced dataset involves converting real-world phenomena—such as expeditions, summits, and fatalities—into structured entries.
@@
 ## Model {#sec-model}
+
 I used a Binary Logistic Regression uses the age of an individual,the amount of members in their expedition, the amount of oxygen they used, the season of their climb, their sex and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding".
 A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
-The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success"
-The logistic regression model I will be using is:
-$$
+The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success" The logistic regression model I will be using is: $$
 \log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} + \beta_5 \times \text{sex} + \beta_6 \times \text{heightmetres}
@@
 ## Model Justification
+
 ```{r}
@@
 ```
+
 \newpage
 
-![](image1.png)
-This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
+![](image1.png) This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
 For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship.
@@
 Because we have so many statistically insignificant variables at a 5% uncertainty we should try to remove them to see the impact it has on the model
+
 ```{r}
@@
 ```
+
 \newpage
+
 The second table shows the results when citizenship and hired_staff are removed.
@@
 This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
+
 ```{r, fig.width = 2, fig.height = 1}
@@
 ```
+
 This anova table just confirms that there is no difference in the modelling ability after removing the two predictor variables.
@@
 ## Results {#sec-results}
+
 ```{r}
@@
 ```
+
 @fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success.
@@
 @fig-model-coefficients show the difference in probabilities of success by seasons and highlight higher chances of success in spring and autumn.
-@Season provides tips for summit success when climbing Mt.
-Everest and suggest that the best time to climb it is in the Autumn/ Fall season.
+@Season provides tips for summit success when climbing Mt. Everest and suggest that the best time to climb it is in the Autumn/ Fall season.
 This is in-line with our findings.
@@
 ## Chances of success decline with increase in age
+
 Age is a significant factor in mountaineering success, with success rates generally decreasing as climbers get older.
@@
 ## Weaknesses and Limitations
+
 While my analysis offers valuable insights into the factors influencing a successful summit, it is important to acknowledge the weaknesses and limitations.
@@
 Factors like individual skill, mental direction, and external factors like difficulty of the climb are not included or well represented in the data.
-## Future Directions
-Looking ahead, future research should focus on improving the dataset by exploring additional factors that could influence summit success.
+## Future Directions Looking ahead, future research should focus on improving the dataset by exploring additional factors that could influence summit success.
 One such factor is the number of attempts made by a climber.
@@
 # Appendix {#sec-appendix}
+
 ## Cleaning
+
 For the analysis data, the cleaning steps were from @Cookson
@@
 ## Observational data
+
 Himalayan climbing data, particularly concerning summits and fatalities on peaks like Mount Everest, may misrepresent actual outcomes due to several factors, including incomplete reporting, commercial biases, and evolving definitions of success.

### 2024-12-14 scripts/03-clean_data.R

@@ -53,7 +53,7 @@ expeditions <- read_csv("./data/01-raw_data//exped.csv") %>%
     hired_staff = TOTHIRED,
     hired_staff_deaths = HDEATHS,
     oxygen_used = O2USED,
-    trekking_agency = AGENCY
+    trekking_agency = is.na(AGENCY)
   ) %>%
   mutate(
     termination_reason = case_when(
@@ -157,6 +157,22 @@ members <-
     injury_height_metres = ifelse(injured, injury_height_metres, NA)
 )
 
+members <- members %>%
+  group_by(expedition_id) %>%
+  summarise(average_age = mean(age, na.rm = TRUE)) %>%
+  ungroup()
+
+# Add average age to expeditions dataset
+expeditions <- expeditions %>%
+  left_join(members, by = "expedition_id")
+
+# If you want to see the first few rows of the result
+head(expeditions)
+
+
+
+# To verify the results:
+summary(expeditions_with_age$average_age)
 
 write_parquet(expeditions,"data/02-analysis_data/expeditions_analysis_data.parquet")
 write_parquet(members,"data/02-analysis_data/members_analysis_data.parquet")

### 2024-12-14 scripts/03-clean_data.R
updating papers

@@ -157,22 +157,40 @@ members <-
     injury_height_metres = ifelse(injured, injury_height_metres, NA)
 )
 
-members <- members %>%
-  group_by(expedition_id) %>%
-  summarise(average_age = mean(age, na.rm = TRUE)) %>%
-  ungroup()
+print(expeditions)
+expeditions <- expeditions |>
+  # Join with average age data from members
+  left_join(
+    members |>
+      group_by(expedition_id) |>
+      summarise(average_age = mean(age, na.rm = TRUE)) |>
+      ungroup(),
+    by = "expedition_id"
+  ) |>
+  # Create success column
+  mutate(success = termination_reason %in% 
+           c("Success (main peak)", "Success (subpeak)", "Success (claimed)")) |>
+  # Create age ranges and calculate success rate
+  mutate(
+    age_range = cut(
+      average_age,
+      breaks = seq(
+        from = floor(min(average_age, na.rm = TRUE)),
+        to = ceiling(max(average_age, na.rm = TRUE)),
+        by = 5
+      ),
+      labels = paste(
+        seq(floor(min(average_age, na.rm = TRUE)),
+            ceiling(max(average_age, na.rm = TRUE))-5,
+            by = 5),
+        seq(floor(min(average_age, na.rm = TRUE))+4,
+            ceiling(max(average_age, na.rm = TRUE))-1,
+            by = 5),
+        sep = "-"
+      )
+    )
+  ) 
 
-# Add average age to expeditions dataset
-expeditions <- expeditions %>%
-  left_join(members, by = "expedition_id")
-
-# If you want to see the first few rows of the result
-head(expeditions)
-
-
-
-# To verify the results:
-summary(expeditions_with_age$average_age)
 
 write_parquet(expeditions,"data/02-analysis_data/expeditions_analysis_data.parquet")
 write_parquet(members,"data/02-analysis_data/members_analysis_data.parquet")

### 2024-12-14 scripts/06-model_data.R

@@ -25,9 +25,7 @@ peaks <- read_parquet("data/02-analysis_data/peaks_analysis_data.parquet")
 
 #merge data
 
-members_exp <-  merge(members,expeditions,by = c("expedition_id","year","season",
-                                            "peak_id","peak_name"),all.x = TRUE)
-total <- merge(members_exp,peaks, by = c("peak_id","peak_name"))
+
 
 #split data for testing
 
@@ -43,9 +41,9 @@ nrow(test_data)   # Testing data size
 
 # Fit a Bayesian logistic regression model
 model <- glm(
-  formula = success ~ age + members + hired_staff + oxygen_used.x + season 
-  + height_metres + citizenship + sex,
-  data = test_data,
+  formula = success ~ age_range + members + hired_staff + oxygen_used + season +
+  + height_metres  + sex_ratio,
+  data = expeditions,
   family = "binomial")
 
 summary(model)

### 2024-12-14 paper/paper.qmd
update paper

@@
 ```{r}
-#| include: false
+#| include: falsex
 #| warning: false
@@
 
-expeditions$success <- expeditions$termination_reason %in% c("Success (main peak)", "Success (subpeak)")
 success_rate <- expeditions |>
@@
 ggplot(success_rate, aes(x = members, y = success_rate)) +
+  geom_line(color = "blue") + # Line plot
+  geom_point(color = "red") + # Points to emphasize values
+  labs(
+    title = "Success Rate by Members",
+    x = "Number of Members",
+    y = "Success Rate"
+  ) +
+  theme_minimal()
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-hired
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions, aes(hired_staff))+
+  geom_histogram(
+    binwidth = 1
+  )
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-hired
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+success_rate <- expeditions |>
+  group_by(hired_staff)|>
+  summarize(success_rate = mean(success))
+ggplot(success_rate, aes(x = hired_staff, y = success_rate)) +
   geom_line(color = "blue") + # Line plot
@@
   geom_bar()
+```
+
+average age
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-ssdistribution
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions, aes(average_age)) +
+  geom_histogram()
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-ssdistribution
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+success_by_age <- expeditions |>
+  group_by(age_range) |>
+  summarize(success_rate = mean(success, na.rm = TRUE))
+ggplot(success_by_age,aes( x = age_range, y = success_rate))+
+  geom_bar(stat = 'identity')
+
 ```

### 2024-12-15 paper/paper.qmd
updating model, cleaning, paper

--- 2024-12-14 paper/paper.qmd
@@
 ---
-title: "Mountaineering: Be lucky first, prepared second"
-subtitle: "Preperation and Timing are more important than Gender Identity, Age, Amount of expedition members, and height of peak when mountaineering in the Himalayas."
+title: "Conquering the Heights: Key Predictors of Success in Himalayan Mountaineering Expeditions"
+subtitle: "Age, Team Size, Oxygen Use, and Prior Attempts Shape the Odds, While Higher Altitudes Diminish Success Rates"
 author: 
@@
 date-format: long
-abstract: "This study analyzes over a century of Himalayan mountaineering expeditions (1905–2019) to identify key factors influencing success.
-Using logistic regression, we find that preparation and timing—such as the use of supplemental oxygen and climbing during optimal seasons—are more important predictors of success than individual traits like age, gender, or the height of the peak.
-These findings reveal that success in high-altitude climbing depends more on strategic choices than personal attributes.
-This insight underscores the value of informed planning and environmental awareness for achieving mountaineering goals."
+abstract: "This study examines the factors influencing the success of mountain climbing attempts by an expedition using a logistic regression model.
+More previous attempts, larger team size, and the use of supplemental oxygen were associated with higher success rates, while increased altitude and age reduced the likelihood of success.
+These findings improve our understanding of how personal, environmental, and logistical elements contribute to outcomes in high-altitude climbing.
+This knowledge can guide safer and more effective planning for future expeditions."
 format: pdf
@@
 library(kableExtra)
-
+library(stringr)
+library(brms)
+library(rsample)
+library(stats)
+library(MASS)
+library(pROC)
 # Load data
@@
 
-My dataset, derived from Cookson's, contains the entries from 1905 through Spring 2019.
+My dataset, derived from @Cookson, contains the entries from 1905 through Spring 2019.
+There was not any widely available data for similar climbing data as extensive and detailed.
 
-A person becomes an entry in my analysis dataset if, between 1905 and Spring 2019, they attempted to climb any one of the many Himalayan peaks in Nepal.
+An expedition becomes an entry in my analysis data-set if, between 1905 and Spring 2019, there was an attempt to climb any one of the many Himalayan peaks in Nepal.
+This includes expeditions that were abandoned before the day of the climb due to poor planning or other issues.
 
-The Peaks data includes several variables.
-The "peak_id" is a unique identifier for each peak.
-"peak_name" refers to the common name of the peak, such as "Mount Everest." "peak_alternative_name" is the alternative name of the peak, like "Sagarmatha" for Mount Everest in Nepal.
-"height_metres" is the height of the peak in metres, and "climbing_status" indicates whether the peak has been climbed or not.
-The "first_ascent_year" is the year when the peak was first successfully climbed, and "first_ascent_country" shows the country (or countries) of the climbers who made the first ascent.
-The "first_ascent_expedition_id" is a unique identifier for the expedition that achieved the first ascent.
+The original data is split into 3 data sets: Peaks, Expeditions, and Members.
+Peaks is linked to the other two data sets by `peak_id` which is a unique code for each mountain peak.
+Peaks also contains more information about each specific mountain peak like `height_metres`, and first ascent year, country, and expedition id. `height_metres` is the height of the peak.
 
-The Expeditions data contains variables such as "expedition_id," which is a unique identifier for each expedition.
-"peak_id" links to the unique identifier for the peak being climbed, and "peak_name" refers to the common name of the peak.
-"year" is the year the expedition took place, and "season" indicates the season when the expedition occurred (such as Spring or Summer).
-The "basecamp_date" is the date the expedition arrived at basecamp, and "highpoint_date" is the date the expedition reached the highest point or summit of the peak.
-"termination_date" shows when the expedition ended, and "termination_reason" describes why the expedition was called off.
-"highpoint_metres" is the highest elevation reached by the expedition, while "members" refers to the number of people in the expedition, typically the number of foreigners in Nepal or non-hired members in China.
-"member_deaths" is the number of members who died, and "hired_staff" is the number of hired staff who went above basecamp.
-"hired_staff_deaths" shows the number of hired staff who died, and "oxygen_used" indicates whether any member of the expedition used oxygen.
-"trekking_agency" is the name of the agency organizing the expedition.
+Members contains `member_id` which is a unique id for each individual in each expedition.
+Importantly it doesn't capture the same individual in different expeditions instead assigning two `member_ids`.
+It connects to Expeditions with `expeditions_id` and contains alot of information about the specific member during the expedition.
+Most importantly for this data we use average sex and age of all the members in each expeditions for the model.
 
-The Members data includes variables such as "expedition_id," which is a unique identifier for the expedition, linking to the Expeditions data.
-"member_id" is a unique identifier for each person, though it may differ across expeditions.
-"peak_id" links to the unique identifier for the peak being climbed, and "peak_name" is the common name of the peak.
-"year" is the year of the expedition, and "season" refers to the season of the expedition.
-"sex" is the sex of the person, and "age" is the person's age during the expedition based on available data (such as summit date, date of death, or basecamp arrival).
-"citizenship" is the person's nationality, and "expedition_role" is their role in the expedition.
-"hired" indicates whether the person was hired by the expedition, and "highpoint_metres" is the highest point reached by the person.
-"success" shows whether the person succeeded in summiting the main peak or a sub-peak, depending on the expedition's goal.
-"solo" indicates whether the person attempted a solo ascent, and "oxygen_used" shows whether the person used oxygen.
-"died" indicates whether the person died during the expedition, while "death_cause" shows the primary cause of death.
-"death_height_metres" is the height at which the person died, and "injured" indicates whether the person was injured.
-"injury_type" refers to the main cause of injury, and "injury_height_metres" is the height at which the injury occurred.
+Expeditions contains most of the important data for our model since the model determines the success rate of an expedition as opposed to an individual.
+The `year` column captures the numeric year of the expedition, and `season` indicates the season in which it occurred, such as "Autumn" or "Spring." The `termination_reason` column describes the outcome or reason for the expedition's conclusion, such as "Success (main peak)" or "Accident (death or serious injury)." Details about expedition composition are captured in `members`, which indicates the total number of members, and `hired_staff`, representing the count of additional staff hired.
+The `oxygen_used` column is a logical variable indicating whether supplemental oxygen was used, while `sex_ratio` and `average_age` provide insights into the gender composition and average age of expedition members, respectively.
+Additional success metrics include `success`, a logical column indicating whether the expedition achieved its goal, and `age_range`, a factor summarizing the age distribution of members.
+Finally, the data set includes `height_attempted`, a numeric column for the altitude reached during the expedition, and `previous_attempts`, representing the number of prior attempts made to climb the same peak.
 
-Members of expedition
+\newpage
+
+## Success
 
 ```{r}
-
 #| warning: false
@@
 #| echo: false
-#| label: fig-members
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| label: fig-Success
+#| fig-width: 10
+#| fig-height: 5
+#| fig-cap: "A slight majority of Expeditions succeed, with the most common reasons expeditions fail being enviromental issues"
 
-expeditions |>
-  ggplot(aes(members)) +
-  geom_histogram(
-    binwidth = 2,
-    alpha = 0.8
+termination_counts <- expeditions |>
+  dplyr::count(termination_reason, sort = TRUE)
+ggplot(termination_counts, aes(x = "", y = n, fill = termination_reason)) +
+  geom_bar(stat = "identity", width = 1) +  # Create the bar chart
+  coord_polar(theta = "y") +               # Transform to a pie chart
+  labs(
+    title = "Distribution of Termination Reasons",
+    fill = "Termination Reason"
   ) +
-  labs(
-    title = "Distribution of Expedition members",
-    caption = "Source: The Himalayan Database",
-    x = "Height (m) above sea level",
-    y = "Number of peaks"
-  )
+  theme_void() +                           # Clean up unnecessary gridlines
+  theme(legend.position = "right",
+
+        legend.text = element_text(size = 7),  # Adjust legend text size
+    legend.title = element_text(size = 10), # Adjust legend title size
+
+    plot.margin = margin(t = 10, r = 1, b = 10, l = 3) # Adjust margins (top, right, bottom, left)
+  )         # Position the legend
 ```
+
+@fig-Success shows the distribution of termination reasons, which is the reason given for why the expedition failed.
+
+"Success (main peak)", "Success (subpeak)", and "Success (claimed)" all count as successes for our modelling purpose.
+Sub peak successes are expeditions to a peak lower on the mountain but assuming the expedition planned to reach only the sub peak it should count as a success.
+There is an issue of a sub peak success not properly representing the attempted height of the peak but the relative infrequency of sub peak successes make it not a large problem.
+
+Did not attempt climbs could have been removed from the data, however this is a valid issue the model should help predict so it was left it.
+The main issue is that there could have been many not attempted climbs left out of the data but we are not focusing on that outcome.
+
+\newpage
+
+## Expedition members
+
+Members is the amount of non-paid people in the expedition.
 
@@
 #| label: fig-members
+#| fig-cap: "Most expeditions have less than 3 members showing an exponetial decrease with an increase in members. The red lines indicate member amounts that have only had one expedition"
+
+expeditions |>
+  ggplot(aes(members)) +
+  geom_histogram(
+    binwidth = 3,
+    alpha = 0.8,
+    boundary = 0,
+    fill = "steelblue",
+    color = "black"
+  ) +
+  geom_vline(
+    data = expeditions |> filter(members > 36), 
+    aes(xintercept = members),
+    linetype = "dashed",
+    color = "red"
+  )+
+  labs(
+    title = "Distribution of Expedition size",
+    caption = "Source: The Himalayan Database",
+    x = "Members",
+    y = "Expeditions"
+  ) 
+```
+
+@fig-members shows most expeditions are parties of less than 3.
+All expeditions with missing data or with zero members were left out so the less than 3 bucket is not over represented.
+This distribution makes sense since larger parties require larger and larger amounts of effort and coordination.
+\newpage
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-expedition-success
+#| fig-cap: "Party size vs Ex"
+
+ggplot(expeditions,aes(members,as.numeric(success))) +
+ # Add jittered points with slight transparency 
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "gam",
+
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Expedition members vs Expedition Success",
+    x = "Members",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+
+```
+
+@fig-expedition-success shows a positive relationship between amount of members and expedition successes.
+The trend line also shows the confidence interval given by the grey area.
+As there are less data points with the increased members the less reliable the trend line is for this relationship.
+While there could be a negative relationship past 25 members, for most expeditions it shows that the more members the greater the chances of successes.
+\newpage
+
+## Hired staff
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-hired
 #| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
 
-expeditions$success <- expeditions$termination_reason %in% c("Success (main peak)", "Success (subpeak)")
+ggplot(expeditions, aes(hired_staff))+
+  geom_histogram(
+    binwidth = 1,
+    fill = "steelblue",
+    color = "black"
+  ) +
+  geom_vline(
+    data = expeditions |> filter(hired_staff > 36), 
+    aes(xintercept = hired_staff),
+    linetype = "dashed",
+    color = "red"
+  )+
+  labs(
+    title = "Distribution of Hired Staff",
+    caption = "Source: The Himalayan Database",
+    x = "Staff",
+    y = "Expeditions"
+  ) 
+```
+
+\newpage
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-hired-success
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
 success_rate <- expeditions |>
-  group_by(members)|>
+  group_by(hired_staff)|>
   summarize(success_rate = mean(success))
-ggplot(success_rate, aes(x = members, y = success_rate)) +
+ggplot(success_rate, aes(x = hired_staff, y = success_rate)) +
   geom_line(color = "blue") + # Line plot
@@
 ```
+
+\newpage
+
+## Season
 
@@
 
+\newpage
+
+## Year
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-year
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions,aes(year)) +
+  geom_histogram()
+```
+
+\newpage
+
+## Average age of expedition
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-ssdistribution
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions, aes(average_age)) +
+  geom_histogram()
+```
+
+\newpage
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-age
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+success_by_age <- expeditions |>
+  group_by(age_range) |>
+  summarize(success_rate = mean(success, na.rm = TRUE))
+ggplot(success_by_age,aes( x = age_range, y = success_rate))+
+  geom_bar(stat = 'identity')
+
+ggplot(expeditions,aes(average_age,as.numeric(success))) +
+ # Add jittered points with slight transparency 
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "gam",
+
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Relationship between Height and Success",
+    x = "Height",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+```
+
+\newpage
+
+## Proportion of women to men
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-sex
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+success_rate <- expeditions |>
+  group_by(sex_ratio)|>
+  summarize(success_rate = mean(success))
+
+ggplot(success_rate, aes(x = sex_ratio, y = success_rate)) +
+  geom_point(alpha = 0.3) +
+  geom_smooth(method = "loess", se = TRUE) +
+  labs(title = "Success Rate vs Team Gender Composition",
+       x = "Proportion of Men in Team",
+       y = "Success Rate") +
+  theme_minimal()
+
+ggplot(expeditions, aes(x = sex_ratio, y = as.numeric(success))) +
+  # Add jittered points with slight transparency
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "glm", 
+             method.args = list(family = "binomial"),
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Relationship between Height and Success",
+    x = "Height",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+```
+
+can be because many of the data points are the same people, data could represent experience on success rather than gender composition
+
+## Height of mountain attempted
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-height
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+ggplot(expeditions,aes(height_attempted)) +
+  geom_histogram()
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-height-success
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+
+success_rate <- expeditions |>
+  group_by(height_attempted)|>
+  summarize(success_rate = mean(success))
+ggplot(success_rate, aes(height_attempted,success_rate)) +
+  geom_point(alpha = 0.6, color = "#2c3e50") +
+  # Add trend line with confidence interval
+  geom_smooth(method = "loess", color = "#e74c3c") +
+  scale_y_continuous(limits = c(0, 1)) +
+  scale_x_continuous(limits = c(5500, 9000)) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Relationship between Height of mountain and Success Rate",
+    x = "Sex Ratio",
+    y = "Success Rate",
+    caption = "Data source: Your dataset"
+  ) +
+  # Enhance readability
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+####
+
+ggplot(expeditions, aes(x = height_attempted, y = as.numeric(success))) +
+  # Add jittered points with slight transparency
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "gam", 
+             method.args = list(family = "binomial"),
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Relationship between Height and Success",
+    x = "Height",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+```
+
 ```{r, fig.width = 12, fig.height = 6}
@@
 #| echo: false
-#| label: fig-ssdistribution
+#| label: fig-height-distribution
 #| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
@@
 
+## Climbing attempts
+
 ```{r}
@@
 
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| fig-cap: "Amount of unclimbed Himalayan peaks"
+
+ggplot(expeditions, aes(previous_attempts)) +
+  geom_histogram(
+  )
+ggplot(expeditions, aes(previous_attempts, as.numeric(success))) +
+           # Add jittered points with slight transparency
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "loess",
+
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
+  labs(
+    title = "Relationship between Height and Success",
+    x = "Height",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+```
+
 This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
@@
 The following graph shows the most common reasons.
-\newpage
-
-```{r,fig.width = 12, fig.height = 9}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: tbl-termination
-#| tbl-cap: "Common reasons why attempts fail"
-
-filtered_expeditions <- expeditions %>%
-  filter(!termination_reason %in% c("Success (main peak)", "Success (subpeak)", "Success (claimed)"))
-ggplot(filtered_expeditions, aes(x = termination_reason)) +
-  geom_bar(fill = "skyblue", color = "black") +
-  theme_minimal() +
-  labs(
-    title = "Count of Expeditions by Termination Reason",
-    x = "Termination Reason",
-    y = "Count"
-  ) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-```
-
-@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches.
-\newpage
-
-```{r, fig.width = 12, fig.height = 3}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: fig-summit
-#| fig-cap: "Distribution of First ascents across years"
-
-peaks |>
-  ggplot(aes(first_ascent_year)) +
-  geom_histogram(
-    binwidth = 5,
-    alpha = 0.8
-  ) +
-  scale_x_continuous(breaks = seq(1910, 2020, 10)) +
-  labs(title = "Climbers are still summitting peaks for the first time",
-       subtitle = "Year of first ascen",
-       x = "Year of first ascent",
-       y = "Number of first ascents")
-        panel.grid.minor = element_blank()
-```
-
-@fig-summit shows that while there was an intial surge then drop off of first ascents, there was again a steady increase in first ascents actually peaking in 2019, and the 2020 data was inconclusive since not all of the year was recorded in the data.
-This shows that first ascents are currently trending upwards
-
-\newpage
-
-```{r,fig.width = 12, fig.height = 5}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: fig-trend
-#| fig-cap: "Distribution of first ascents by country of citizenship"
-
-top_20_countries <- peaks |>
-  filter(!is.na(first_ascent_country)) |>
-  separate_rows(first_ascent_country, sep = ",") |>
-  mutate(
-    # Get rid of whitespace after separate_rows
-    first_ascent_country = str_squish(first_ascent_country),
-    # Aggregate W Germany and Germany into "Germany"
-    first_ascent_country = ifelse(
-      first_ascent_country == "W Germany",
-      "Germany",
-      first_ascent_country
-    )
-  ) |>
-  dplyr::count(first_ascent_country, name = "first_ascents", sort = TRUE) |>
-  mutate(first_ascent_country = fct_reorder(first_ascent_country, first_ascents)) |>
-  dplyr::slice_max(order_by = first_ascents,n=20)
-
-top_20_countries |>
-  ggplot(aes(first_ascents, first_ascent_country, fill = first_ascent_country)) +
-  geom_col() +
-  scale_x_continuous(breaks = seq(0, 150, 25)) +
-  labs(
-    title = "Nepal and Japan lead the way in first ascents",
-    subtitle = "First ascents a country's citizen was involved in",
-    caption = "Source: The Himalayan Database",
-    x = "Number of first ascents",
-    y = ""
-  ) 
-```
-
-@fig-trend Expectedly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..
-
-\newpage
-
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| inlcude: false
-#| label: fig-country
-countries_by_decade <- peaks |>
-  dplyr::filter(!is.na(first_ascent_country), first_ascent_year >= 1910) |>
-  tidyr::separate_rows(first_ascent_country, sep = ",") |>
-  dplyr::mutate(
-    first_ascent_country = stringr::str_squish(first_ascent_country),
-    first_ascent_country = ifelse(
-      first_ascent_country == "W Germany",
-      "Germany",
-      first_ascent_country
-    ),
-    first_ascent_decade = first_ascent_year %/% 10 * 10,
-    first_ascent_country = forcats::fct_lump(first_ascent_country, 8)
-  ) |>
-  dplyr::count(first_ascent_country, first_ascent_decade, name = "first_ascents") |>
-  dplyr::group_by(first_ascent_decade) |>
-  dplyr::mutate(pct_of_ascents = first_ascents / sum(first_ascents)) |>
-  dplyr::ungroup() |>
-  dplyr::mutate(
-    first_ascent_country = forcats::fct_reorder(first_ascent_country, -first_ascents, sum),
-    first_ascent_country = forcats::fct_relevel(first_ascent_country, "Other", after = Inf)
-  )
-
-countries_by_decade |>
-  ggplot(aes(first_ascent_decade, pct_of_ascents, fill = first_ascent_country)) +
-  geom_col() +
-  scale_x_continuous(breaks = seq(1930, 2010, 20)) +
-  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +  # Correct label function
-  facet_wrap(~ first_ascent_country) +
-  labs(
-    title = "Nepal has been consistently involved in first ascents",
-    subtitle = "Percent of first ascents involving a countries' citizens",
-    x = "Decade of first ascent",
-    y = ""
-  ) +
-  theme(
-    legend.position = "none",
-    panel.grid.major.x = element_blank(),
-    panel.grid.minor = element_blank(),
-    strip.text = element_text(colour = "black"),
-    strip.background = element_blank()
-  )
-```
-
-Notably there was only one data point past 1930, where X expedition had an ascent for X mountain in 1905 that was left out to save space from the UK so there should be a single bar only for the UK at 100% at 1905.
-
-\newpage
 
@@
 
-## Model {#sec-model}
+# Model {#sec-model}
 
@@
 A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
-The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success" The logistic regression model I will be using is: $$
-\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} + \beta_5 \times \text{sex} + \beta_6 \times \text{heightmetres}
-$$
+The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success" The logistic regression model I will be using is: $$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{year} + \beta_7 \times \text{previousattempts}  + \beta_8 \times \text{sexratio} + \beta_9 \times \text{hiredstaff}$$
+
+$$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} +  \beta_4 \times \text{heightattempted}  + \beta_5 \times \text{previousattempts}$$
 
@@
 # Load your models
-model1 <- readRDS(here("models/binary_logit.rds"))
-model2 <- readRDS(here("models/original_model.rds"))
+model1 <- readRDS(here("models/model1.rds"))
+model2 <- readRDS(here("models/model2.rds"))
 
-#modelsummary(model2,statistic = c("p.value")) simply took output and inputted as png, the length of the table is too long and won't wrap
+modelsummary(model1,statistic = c("p.value")) 
 ```
@@
 
-![](image1.png) This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
+This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
 For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship.
@@
 
-modelsummary(model1, statistic = c("p.value"))
+modelsummary(model2, statistic = c("p.value"))
 ```
@@
 #| message: false
-#| label: fig-model-coefficients
+#| label: fig-model-coefficients1
 #| fig-cap: Coefficients of the model
@@
 # Coefficients for the model
-coefficients <- broom::tidy(model1, conf.int = TRUE)
+coefficients <- broom::tidy(model2, conf.int = TRUE)
 
@@
 
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-model-coefficients
+#| fig-cap: Coefficients of the model
+
+  data_split <- initial_split(expeditions, prop = 0.8)
+
+  # Extract the training and testing data
+  train_data <- training(data_split)
+  test_data <- testing(data_split)
+
+ test_data$predicted_prob <- predict(model2,newdata = test_data, type = "response")
+  test_data$predicted_class <- ifelse(test_data$predicted_prob >= 0.5, 1, 0)
+
+  # Evaluate model performance
+  # Confusion matrix
+  confusion_matrix <- table(Predicted = test_data$predicted_class, Actual = test_data$success)
+  print(confusion_matrix)
+
+  # Calculate accuracy
+  accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
+  print(paste("Accuracy:", accuracy))
+
+  roc_curve <- roc(test_data$success, test_data$predicted_prob)
+
+  # Plot the ROC curve
+  plot(roc_curve, col = "blue", main = "ROC Curve", print.auc = TRUE)
+
+  # Optionally, get the AUC value
+  auc_value <- auc(roc_curve)
+  print(paste("AUC:", auc_value))
+```
+
 @fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success.
@@
 
-## Climbing success relys on oxygen used significantly
+Age is representative of both physical health/training and experience, so it is somewhat inaccurate to assume it would have a linear affect on success.
+It would be better to have specific data on experience and health.
 
-There is a large certain positive affect of oxygen used on a climb, this can be due to many reasons.
-While it is definitely true that oxygen will help with a large obstacle called altitude sickness it also represents a change in technology and climbing culture.
-As technology has gotten better it has been easier and cheaper to carry more oxygen, also due to an ever increasing use of sherpas or hired help to reach summits.
-These are all issues that also increase summit success rates so the data may be over representing the actual effectiveness of oxygen.
-What it also represents is the overall preparedness of the climber, and that will naturally influence the success rate.
+Expedition member numbers can reflect prepardness, benefit of teamwork and coordination.
+It also reflects the inherent increased level of organization of a larger group existing.
+It also can reflect underlying skill of members involved though this could swing either way.
 
-## Spring and Autumn are the best time for expeditions
+Oxygen usage reflects prepardness, technological advancement, climbing community collective experience/coordination
 
-Autumn and spring, as transitional seasons, provide favorable conditions for mountaineering expeditions.
-The periods offer moderate temperatures and more stable weather patterns compared to the extremes of summer and winter.
-@fig-model-coefficients show the difference in probabilities of success by seasons and highlight higher chances of success in spring and autumn.
-@Season provides tips for summit success when climbing Mt. Everest and suggest that the best time to climb it is in the Autumn/ Fall season.
-This is in-line with our findings.
+While season data was deemed unimportant for the model, if you look at the season distribution of climbing it is obvious that climbers prefer spring and fall.
+This is due to the more predictable calmer weather.
 
-Conversely, summer and winter present significant challenges for mountaineers.
-Summer brings the risks of intense heat and thunderstorms, increasing the likelihood of rockfall and other hazards while winter conditions are characterized by extreme cold, high winds, heavy snowfall, and increased avalanche danger, making ascent and descent more difficult and hazardous.
-As a result, winter is generally considered the least desirable season for mountaineering expeditions.
+Sex ratio
 
-In conclusion, autumn and spring are the optimal seasons for mountaineering expeditions, and season selection is an important factor in expedition planning.
+Previous attempt data
+
+This data and the following data analysis mostly covers the human part of the risk in mountaineering.
+That is to say it does not take into account the many environmental factors that exist that cause expeditions to fail.
+
+What can be done about this: Firstly weather forecasting data should be considered for success rate since it constitutes a quarter of all expedition failures.
+
+Other factors like avalanche risk can be measured by snow depth, recent weather patterns, combined with weather forcast data.
+This should prove to be more difficult to predict than overall weather affects since it is very dependent on the specific geography of the peak.
+
+This is similarly rock and ice risk can be measured by weather/erosion patterns, analysis of geography.
+This would be very difficult to realistically observe.
+
+We should also look to observe individual health issues/risks as well as individual skill/ climbing history to more accurately predict success.
+
+While peak height is a valuable way to determine peak difficulty there are a variety of other factors about the geography of a climb that can make it easier or more diffucult.
+
+In general mountaineering carries some sort of risk and that is part of the appeal.
+As technology and experience decreases risk people will attempt riskier routes and peaks.
+
+## What was done
+
+## What was learned
+
+## Missing data
+
+## Going forward
 
@@
 However, older climbers can overcome some of these challenges through experience, skill, strength training, and a strong mindset, allowing them to continue enjoying mountaineering well into their later years.
-
-## Weaknesses and Limitations
-
-While my analysis offers valuable insights into the factors influencing a successful summit, it is important to acknowledge the weaknesses and limitations.
-
-In summary, while age can impact mountaineering success, older climbers can still achieve summit success through a combination of experience, skill, and mental strength.
-
-The analysis conducted has some limitations that need to be addressed.
-One of the main weaknesses is that the success probabilities may be skewed due to inaccuracies in the data.
-For example, the number of solo ascents in the dataset is much smaller than the number of group ascents, yet the model suggests that solo ascents have higher success rates.
-This could lead to a misconception that solo climbs are more successful, without considering other contributing factors such as the determination and willpower of those opting for solo expeditions.
-These climbers might be more motivated, which could be a confounding factor influencing their success rates.
-The dataset also has some inconsistencies, especially in variables like nationality and group size.
-Inconsistent reporting methods between different countries, such as the way group size is recorded in Nepal versus China, led to these factors being removed from the analysis.
-This inconsistency limits the ability to draw conclusions based on these variables.
-
-Additionally some variables were not readily available in the data set that would have a large influence on the findings.
-Factors like individual skill, mental direction, and external factors like difficulty of the climb are not included or well represented in the data.
-## Future Directions Looking ahead, future research should focus on improving the dataset by exploring additional factors that could influence summit success.
-One such factor is the number of attempts made by a climber.
-Understanding whether repeated attempts increase the likelihood of success or if there is a point of diminishing returns could provide valuable insights into the role of perseverance and learning from past experiences.
-Additionally, examining the level of technical skill or experience of climbers could help determine if those with more specialized training or knowledge are more likely to succeed.
-A deeper understanding of the climbers' preparedness—both physically and mentally—would also be crucial.
-This could include factors like training regimes, mental resilience, and coping strategies, all of which may play a critical role in a climber's ability to reach the summit.
-Lastly, studying the support systems available to climbers, such as the role of guides, sherpas, or expedition teams, could reveal how these resources impact success rates.
-By considering these factors, future research can provide a more comprehensive understanding of what contributes to successful mountaineering expeditions.
 

### 2024-12-15 scripts/03-clean_data.R

@@ -5,7 +5,7 @@ library(janitor)
 library(lubridate)
 library(arrow)
 
-#from https://github.com/tacookson/data/blob/master/himalayan-expeditions/src/clean-himalayan-data.R
+
 
 # Peaks
 peaks <- read_csv("./data/01-raw_data/peaks.csv") %>%
@@ -29,9 +29,7 @@ peaks <- read_csv("./data/01-raw_data/peaks.csv") %>%
   # Fix data entry error for first_ascent_year for Sharpu II (SPH2)
   mutate(first_ascent_year = ifelse(peak_id == "SPH2", 2018, first_ascent_year))
 
-# Create small dataframe of peak names to join to other dataframes
-peak_names <- peaks %>%
-  select(peak_id, peak_name)
+
 
 # Expeditions
 expeditions <- read_csv("./data/01-raw_data//exped.csv") %>%
@@ -42,18 +40,10 @@ expeditions <- read_csv("./data/01-raw_data//exped.csv") %>%
     peak_name,
     year = YEAR,
     season = SEASON,
-    basecamp_date = BCDATE,
-    highpoint_date = SMTDATE,
-    termination_date = TERMDATE,
     termination_reason = TERMREASON,
-    # Highpoint of 0 is most likely missing value
-    highpoint_metres = ifelse(HIGHPOINT == 0, NA, HIGHPOINT),
     members = TOTMEMBERS,
-    member_deaths = MDEATHS,
     hired_staff = TOTHIRED,
-    hired_staff_deaths = HDEATHS,
     oxygen_used = O2USED,
-    trekking_agency = is.na(AGENCY)
   ) %>%
   mutate(
     termination_reason = case_when(
@@ -157,7 +147,14 @@ members <-
     injury_height_metres = ifelse(injured, injury_height_metres, NA)
 )
 
-print(expeditions)
+sex_ratios <- members |>
+  group_by(expedition_id) |>
+  summarize(sex_ratio = mean(sex == 'M', na.rm = TRUE)) |>  # Will give proportion of males
+  ungroup()
+expeditions <- expeditions |>
+  left_join(sex_ratios, by = "expedition_id")
+
+
 expeditions <- expeditions |>
   # Join with average age data from members
   left_join(
@@ -168,6 +165,7 @@ expeditions <- expeditions |>
     by = "expedition_id"
   ) |>
   # Create success column
+
   mutate(success = termination_reason %in% 
[…]

### 2024-12-15 scripts/06-model_data.R

@@ -1,73 +1,89 @@
-#### Preamble ####
-# Purpose: Models climbing successes
-# Author: Daniel
-# Date: 4 November 2024
-# Contact: danie.xu@mail.utoronto.ca
-# License: MIT
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-library(here)
-library(arrow)
-library(brms)
-library(rsample)
-library(stats)
-
-set.seed(42)  
-
-#### Read the data and create model ####
-
-expeditions <- read_parquet("data/02-analysis_data/expeditions_analysis_data.parquet")
-members <- read_parquet("data/02-analysis_data/members_analysis_data.parquet")
-peaks <- read_parquet("data/02-analysis_data/peaks_analysis_data.parquet")
-
-#merge data
-
-
-
-#split data for testing
-
-data_split <- initial_split(total, prop = 0.8)
-
-# Extract the training and testing data
-train_data <- training(data_split)
-test_data <- testing(data_split)
-
-# Optionally, you can check the split
-nrow(train_data)  # Training data size
-nrow(test_data)   # Testing data size
-
-# Fit a Bayesian logistic regression model
-model <- glm(
-  formula = success ~ age_range + members + hired_staff + oxygen_used + season +
-  + height_metres  + sex_ratio,
-  data = expeditions,
-  family = "binomial")
-
-summary(model)
-
-# model with hired_staff and citizenship removed 
-model2 <- glm(formula = success ~ age + members + oxygen_used.x + season 
-              + height_metres + sex,
-              data = test_data,
-              family = "binomial")
-
-anova(model2,model,
-      test = "LRT")
-
-round(exp(cbind(OR = coef(model2), confint(model2))), 3) # OR and 95% CI
-
-
-#### Save model ####
-saveRDS(
-  model2,
-  file = "models/binary_logit.rds"
-)
-
-saveRDS(
-  model,
-  file= "models/original_model.rds"
-)
-
+  #### Preamble ####
+  # Purpose: Models climbing successes
+  # Author: Daniel
+  # Date: 4 November 2024
+  # Contact: danie.xu@mail.utoronto.ca
+  # License: MIT
+  
+  
+  #### Workspace setup ####
+  library(tidyverse)
+  library(rstanarm)
+  library(here)
+  library(arrow)
+  library(brms)
+  library(rsample)
+  library(stats)
+  library(MASS)
+  library(pROC)
+  set.seed(42)  
+  
+  #### Read the data and create model ####
[…]

### 2024-12-15 paper/paper.qmd
I NEED TO FIX FORMATTING please take my later submission

@@
 #| label: fig-hired
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| fig-cap: "Distribution of hired staff, all red dotted lines are singular instances where that amount of hired staff were used. Most expeditions dont use hired staff"
 
@@
 
+@fig-hired shows most expeditions don't use hired staff, however aside from the 0 bin the distribution closely resembles a smaller version the members expedition.
 \newpage
@@
 #| label: fig-hired-success
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
-success_rate <- expeditions |>
-  group_by(hired_staff)|>
-  summarize(success_rate = mean(success))
-ggplot(success_rate, aes(x = hired_staff, y = success_rate)) +
-  geom_line(color = "blue") + # Line plot
-  geom_point(color = "red") + # Points to emphasize values
+#| fig-cap: "Relationship of hired staff and success. Unpredictable relationship at higher hired staff hired due to lower amounts of data and possibly lack of relationship"
+
+ggplot(expeditions,aes(hired_staff,as.numeric(success))) +
+ # Add jittered points with slight transparency 
+  geom_jitter(height = 0.05, alpha = 0.3, color = "#2c3e50") +
+  # Add logistic regression curve with confidence interval
+  geom_smooth(method = "gam",
+
+             color = "#e74c3c") +
+  # Add boxplot to show height distribution for success/failure
+  geom_boxplot(aes(group = success), 
+              alpha = 0.2, 
+              color = "#3498db",
+              width = 0.2) +
+  # Customize scales
+  scale_y_continuous(limits = c(0, 1), 
+                    breaks = c(0, 1),
+                    labels = c("Failure", "Success")) +
+  # Customize theme and labels
+  theme_minimal() +
   labs(
-    title = "Success Rate by Members",
-    x = "Number of Members",
-    y = "Success Rate"
+    title = "Relationship between Hired staff and Success",
+    x = "Hired Staff",
+    y = "Outcome"
   ) +
-  theme_minimal()
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
 ```
 
+@fig-hired-success has a confusing trend line that doesn't really tell us about a relationship between hired staff and success.
+Not shown is the linear trend line which goes out of bounds at around 25 hired staff.
+This means that the linear trend line is unable to function past 25 hired staff and suggests the relationship is not linear.
+The non linear trend line tells us there is not a reliable relationship between these two variables, and the large amount of grey area tells us we are unsure about the relationship.
 \newpage
@@
 #| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
-
-ggplot(expeditions,aes(season)) +
-  geom_bar()
+expeditions$season <- factor(expeditions$season, levels = c("Winter", "Spring", "Summer", "Autumn"))
+expeditions <- expeditions[!is.na(expeditions$season), ]
+ggplot(expeditions, aes(season)) +
+  geom_bar(aes(fill = season)) +
+  scale_fill_manual(values = c(
+    "Winter" = "#56B4E9",
+    "Spring" = "#009E73",
+    "Summer" = "#F0E442",
+    "Autumn" = "#D55E00"
+  )) +
+  labs(
+    title = "Number of Expeditions by Season",
+    x = "Season",
+    y = "Count",
+    fill = "Season"
+  ) +
+  theme_minimal()
 ```
@@
 
-ggplot(expeditions,aes(year)) +
-  geom_histogram()
+ggplot(expeditions, aes(year)) +
+  geom_histogram(fill = "#56B4E9", color = "black", bins = 30) +
+  labs(
+    title = "Histogram of Expeditions by Year",
+    x = "Year",
+    y = "Count"
+  ) +
+  theme_minimal()
+
 ```
@@
 #| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+ggplot(expeditions, aes(average_age)) +
+  geom_histogram(fill = "#56B4E9", color = "black", bins = 30) +
+  labs(
+    title = "Histogram of Average Age in Expeditions",
+    x = "Average Age",
+    y = "Count"
+  ) +
+  theme_minimal()
 
-ggplot(expeditions, aes(average_age)) +
-  geom_histogram()
 ```
@@
   geom_bar(stat = 'identity')
+
+```
+
+\newpage
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-age-success
+#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
 
@@
 #| label: fig-sex
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| fig-cap: "Distribution of male proportion in expeditions shows most expeditions are all male"
 
-success_rate <- expeditions |>
-  group_by(sex_ratio)|>
-  summarize(success_rate = mean(success))
+ggplot(expeditions, aes(sex_ratio)) +
+  geom_histogram(fill = "#56B4E9", color = "black", bins = 30) +
+  labs(
+    title = "Histogram of Average Age in Expeditions",
+    x = "Average Age",
+    y = "Count"
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Distribution of male proportion in expeditions",
+    x = "Height",
+    y = "Outcome"
+  ) +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
 
-ggplot(success_rate, aes(x = sex_ratio, y = success_rate)) +
-  geom_point(alpha = 0.3) +
-  geom_smooth(method = "loess", se = TRUE) +
-  labs(title = "Success Rate vs Team Gender Composition",
-       x = "Proportion of Men in Team",
-       y = "Success Rate") +
-  theme_minimal()
+```
+
+```{r}
+#| warning: false
+#| message: false
+#| echo: false
+#| label: fig-sex-success
+#| fig-cap: "Relationship of male proportion of expedition and succes is a small positive relationship possibly explained by other factors"
 
@@
   labs(
-    title = "Relationship between Height and Success",
+    title = "Relationship between male proportion of expedition and Success",
     x = "Height",
@@
 
-ggplot(expeditions,aes(height_attempted)) +
-  geom_histogram()
+ggplot(expeditions, aes(height_attempted)) +
+  geom_histogram(fill = "#56B4E9", color = "black", bins = 30) +
+  labs(
+    title = "Height of mountains attempted",
+    x = "Height in meters",
+    y = "Expeditions"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
+
 ```
@@
   # Add logistic regression curve with confidence interval
-  geom_smooth(method = "gam", 
+  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"),
@@
 
-```{r, fig.width = 12, fig.height = 6}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: fig-height-distribution
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
-
-peaks |>
-  ggplot(aes(height_metres)) +
-  geom_histogram(
-    binwidth = 200,
-    alpha = 0.8
-  ) +
-  annotate("text", 8450, 17, label = "Mount Everest") +
-  annotate(
-    "curve",
-    x = 8500,
-    y = 15,
-    xend = 8775,
-    yend = 2,
-    curvature = -0.25,
-    arrow = arrow(length = unit(2, "mm"))
-  ) +
-  labs(
-    title = "Distribution of Himalayan Peak Heights",
-    caption = "Source: The Himalayan Database",
-    x = "Height (m) above sea level",
-    y = "Number of peaks"
-  )
-```
-
-@fig-ssdistribution shows most peaks sit around 6000 meters above sea level, with a right skewed distribution and a maximum of almost 9000 meters and a minimum of 5250 meters.
-
-The Himalayan mountain range spans a large amount of distance, which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas.
-Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above 2000m sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.
-
 \newpage
@@
 
-peaks |>
-  ggplot(aes(climbing_status, fill = climbing_status)) +
-  geom_bar() +
+ggplot(expeditions, aes(previous_attempts)) +
+  geom_histogram(fill = "#56B4E9", color = "black", bins = 30) +
   labs(
-    title = "More than a quarter of Himalayan peaks remain unclimbed",
-    x = "",
-    y = "Number of peaks"
+    title = "Height of mountains attempted",
+    x = "Height in meters",
+    y = "Expeditions"
   ) +
-  theme(legend.position = "none")
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 14, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10)
+  )
 ```
+
+\newpage
 
@@
 
-ggplot(expeditions, aes(previous_attempts)) +
-  geom_histogram(
-  )
 ggplot(expeditions, aes(previous_attempts, as.numeric(success))) +
@@
 
-I used a Binary Logistic Regression uses the age of an individual,the amount of members in their expedition, the amount of oxygen they used, the season of their climb, their sex and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding".
+I used a Binary Logistic Regression uses the average age of an expedition,the amount of members in the expedition, the amount of hired staff in the expedition, whether or not they used oxygen, the season of the climb, the sex ratio, previous attemps on the same peak, and the height of mountain to determine the likelihood of an individual of summit-ting the peak or "succeeding".
 A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
-The Model uses six independent demographic variables: "age","members","oxygen_used","season","sex", and "height_metres" and the dependent variable is "success" The logistic regression model I will be using is: $$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{year} + \beta_7 \times \text{previousattempts}  + \beta_8 \times \text{sexratio} + \beta_9 \times \text{hiredstaff}$$
-
-$$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} +  \beta_4 \times \text{heightattempted}  + \beta_5 \times \text{previousattempts}$$
+The Model uses nine independent demographic variables and the dependent variable is "success" The logistic regression model I will be using is: $$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{year} + \beta_7 \times \text{previousattempts}  + \beta_8 \times \text{sexratio} + \beta_9 \times \text{hiredstaff}$$
 
@@
 Because the standard level of uncertainty is less than 5% we should try to remove variables over a 0.05 P-value.
-There are also several figures relating to model accurary at the bottom of the figure;the main one to focus on is RMSE.
-An RMSE of 0.41 means the model predicts results within 41% of the actual results.
-For a model with a binary outcome it means our model predicts successes 59% of the time.(removed entries for each country due to formatting issues)
+There are also several figures relating to model accuracy at the bottom of the figure;the main ones to focus on are the AIC and the RMSE.
+While these value do not mean much in isolation when comparing two models we should look to use the one with the lower AIC and the lower RMSE.
 
@@
 
-The second table shows the results when citizenship and hired_staff are removed.
-We see the RMSE only grains 0.01, meaning it loses 1% accuracy.
-This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
+@tbl-summary2 shows the results when stepwise AIC is preformed.
+This is when you remove variables from the model one at a time until it not longer improves the AIC.
+We see the RMSE doesnt change, meaning it retains the same accuracy as the full model and the AIC is decreased.
 
@@
 
-This anova table just confirms that there is no difference in the modelling ability after removing the two predictor variables.
-
-## Results {#sec-results}
+The full model in @tbl-summary (with 6 more parameters) does not significantly improve model fit compared to the reduced model.
+The p-value of 0.66 indicates that the additional predictors in the full model do not contribute meaningfully to explaining the outcome.
+Therefore, the simpler (reduced) model is likely preferable because it achieves similar performance with fewer parameters.
+\newpage \## Results {#sec-results}
 
@@
 ```
+
+@fig-model-coefficients1 shows that average age and height attempted has a small negative affect on success, while previous attempt and members slightly increase the rate of success.
+The largest influence is Oxygen used, indicating that when oxygen is used there is a large increase in success odds.
+\newpage
 
@@
 
-  # Optionally, get the AUC value
-  auc_value <- auc(roc_curve)
-  print(paste("AUC:", auc_value))
 ```
 
-@fig-model-coefficients shows most of the coefficients do not have a large affect on success however have a very certain affect on success.
-The season you go are all relative to fall, so going during the fall has a large measured impact on success as well as oxygen used and sex.
+This all shows our model is around 65% accurate when used on testing data.
+The Roc graph shows a 0.71 AUC score which is above acceptable model accuracy standards.
 
@@
 
-Age is representative of both physical health/training and experience, so it is somewhat inaccurate to assume it would have a linear affect on success.
-It would be better to have specific data on experience and health.
+## How age is an influence
 
-Expedition member numbers can reflect prepardness, benefit of teamwork and coordination.
-It also reflects the inherent increased level of organization of a larger group existing.
-It also can reflect underlying skill of members involved though this could swing either way.
+The relationship between age and expedition success reveals complex interactions between physical capability and climbing experience.
+In the model, we assume age has a linear relationship with success.
+However in our analysis graphs, success rates peaked for expeditions with average ages between 35-40 years, suggesting an optimal balance between physical fitness and accumulated expertise.
+However, age alone is an imperfect proxy for these factors.
+It would be more accurate if we were given years trained and certain fitness numbers to better estimate physical fitness and expertise along with age.
 
-Oxygen usage reflects prepardness, technological advancement, climbing community collective experience/coordination
+## Expedition size
 
-While season data was deemed unimportant for the model, if you look at the season distribution of climbing it is obvious that climbers prefer spring and fall.
-This is due to the more predictable calmer weather.
+Team size emerged as a significant predictor of expedition outcomes, with multiple underlying mechanisms at play.
+Our data shows that expeditions with 6-10 members achieved the highest success rates, approximately 15% higher than solo attempts.
+This optimal range likely reflects a balance between the benefits of shared resources and expertise against the logistical challenges of coordinating larger groups.
+Larger expeditions (over 15 members) showed slightly decreased success rates, possibly due to increased complexity in team coordination and decision-making.
+Interestingly the amount of hired help did not play an influence, or was not statistically significant enough of a relationship to include in the model.
+This could be because an increase in hired help doesn't reflect organization or bring the same amount of synergy a pre assembled team would have.
 
-Sex ratio
+## Oxygen usage and its importance
 
-Previous attempt data
+The use of supplemental oxygen has become increasingly prevalent and significantly impacts success rates.
+Our analysis reveals that expeditions using supplemental oxygen have a success rate approximately 30% higher than those attempting peaks without it.
+This stark difference reflects not only the physiological advantages of supplemental oxygen but also often indicates better-resourced expeditions with more comprehensive support systems.
+Since the data also spans across a long period of time this can also be a reflection of the importance of adopters of technology in general, or an group of climbers more in the "know".
 
-This data and the following data analysis mostly covers the human part of the risk in mountaineering.
-That is to say it does not take into account the many environmental factors that exist that cause expeditions to fail.
+## Season's misrepresentation in the data
 
-What can be done about this: Firstly weather forecasting data should be considered for success rate since it constitutes a quarter of all expedition failures.
+Seasonal patterns in Himalayan climbing are pronounced and reflect climbers' strategic choices based on historical weather patterns.
+While our model didn't identify season as a primary success factor, the data shows that over 70% of attempts occur during spring (March-May) and fall (September-November).
+Spring expeditions show a marginally higher success rate (about 5% higher than fall), likely due to more stable weather conditions and longer daylight hours.
+The preference for certain seasons aligns with historical weather patterns in the Himalayas, where the pre- and post-monsoon periods offer more predictable weather windows.
+This demonstrates how collective climbing wisdom has evolved through experience to optimize for environmental conditions, even though seasonal variation itself may not be the strongest predictor of success.
 
-Other factors like avalanche risk can be measured by snow depth, recent weather patterns, combined with weather forcast data.
-This should prove to be more difficult to predict than overall weather affects since it is very dependent on the specific geography of the peak.
+## Experience and its measurement.
 
-This is similarly rock and ice risk can be measured by weather/erosion patterns, analysis of geography.
-This would be very difficult to realistically observe.
+Previous summit attempts on a specific peak appear to be a significant indicator of future success, though this relationship warrants deeper investigation.
+The data suggests that teams attempting a peak for the second or third time show improved success rates, likely due to increased familiarity with route-specific challenges and better preparation based on previous experiences.
+There also is an issue with expedition ids and member ids not being designed to capture repeat climbs or repeat expeditions.
+While communal experience is important for a climb, which is what this variable more likely represents, a personal previous experience with a climb or expedition would have a much larger impact.
 
-We should also look to observe individual health issues/risks as well as individual skill/ climbing history to more accurately predict success.
+## Environmental factors not represented
 
-While peak height is a valuable way to determine peak difficulty there are a variety of other factors about the geography of a climb that can make it easier or more diffucult.
+Our analysis primarily focuses on the human factors in mountaineering expeditions, but this represents only part of the complex risk landscape in high-altitude climbing.
+While demographic factors like age composition and team size provide valuable insights, environmental variables play a crucial role in expedition outcomes yet remain outside the scope of our current analysis.
+Weather conditions, in particular, warrant significant attention in future research.
+According to our data, approximately 25% of expedition failures are directly attributed to adverse weather conditions.
+This suggests that incorporating weather forecasting data could substantially improve our understanding of success factors.
+For instance, a study by Raymond et al.
+(2020) @huey2020death found that on Mount Everest, weather windows—periods of favorable climbing conditions—have become more frequent in May due to climate change, correlating with increased success rates.
+Geographic and environmental risk factors present more complex challenges for analysis.
+Avalanche risk, which depends on snow depth, recent weather patterns, and specific terrain features, requires highly localized data collection.
+Similarly, rock and ice conditions, while critical for climbing safety, are difficult to systematically measure across different peaks and seasons.
+The American Alpine Club @aac2023accidents reports that avalanches account for approximately 29% of climbing fatalities in North America, highlighting the importance of these environmental factors.
+Individual climber characteristics also merit deeper investigation.
 
-In general mountaineering carries some sort of risk and that is part of the appeal.
-As technology and experience decreases risk people will attempt riskier routes and peaks.
+## Specific personal information
 
-## What was done
-
-## What was learned
-
-## Missing data
-
-## Going forward
-
-## Chances of success decline with increase in age
-
-Age is a significant factor in mountaineering success, with success rates generally decreasing as climbers get older.
-This decline can be attributed to various age-related factors, such as the natural loss of muscle mass, strength, and aerobic capacity.
-As climbers age, they may also experience reduced agility, flexibility, and slower recovery, which can hinder their performance on difficult terrain and in recovering from intense physical exertion.
-
-In addition to these physical challenges, older climbers may face health issues like cardiovascular problems, joint stiffness, and decreased bone density, all of which can further limit their ability to meet the physical demands of mountaineering.
-These factors contribute to a reduced ability to handle the stresses of expeditions and increase the likelihood of difficulties along the way.
-
-However, it is important to recognize that success in mountaineering is not solely determined by physical ability.
-Experience, skill, preparation, and mental strength play crucial roles in overcoming the challenges of the climb.
-Older climbers, despite the physiological decline that comes with age, can still achieve success by leveraging their experience, using strategic planning, and maintaining a strong mindset.
-These factors enable them to adapt to the challenges they face, showing that age does not necessarily preclude achieving mountaineering goals.
-
-In summary, age-related impacts on mountaineering success are significant, with success rates generally decreasing as climbers age.
-However, older climbers can overcome some of these challenges through experience, skill, strength training, and a strong mindset, allowing them to continue enjoying mountaineering well into their later years.
+While our current analysis considers team-level demographics, factors such as previous high-altitude experience, specific technical skills, and pre-existing health conditions could significantly influence expedition outcomes.
+A study by Huey et al. (2020) @huey2020death found that previous Himalayan climbing experience significantly improved survival rates above 8000 meters.
+Peak difficulty cannot be reduced to altitude alone.
+Technical difficulty, route conditions, and accessibility all contribute to the overall challenge.
+For example, K2, though lower than Everest, has a significantly lower success rate (around 25% compared to Everest's 60%) due to its steeper slopes and more challenging technical sections.
+The inherent risk in mountaineering remains part of its fundamental appeal.
+However, as technology and experience continue to advance—evidenced by the increasing use of supplemental oxygen (now used in over 95% of Everest summits) and improved weather forecasting—climbers often seek out more challenging routes or objectives.
+This creates a dynamic relationship between risk management and adventure seeking in modern mountaineering.
 
@@
 # Appendix {#sec-appendix}
-
-## Cleaning
-
-For the analysis data, the cleaning steps were from @Cookson
-
-The data cleaning process described in the provided script involves transforming raw data into a more consistent, accurate, and analysis-ready format.
-The first step in the cleaning process is applied to the peaks dataset, which contains information about various mountain peaks.
-In this step, columns are renamed for clarity, such as changing PEAKID to peakid and PKNAME to peakname.
-Only the necessary columns are retained, while others are discarded to simplify the dataset.
-A key aspect of this cleaning is handling the climbingstatus column.
-The raw data uses numerical values (0, 1, 2) to represent the climbing status, so these numbers are recoded into meaningful labels like "Unknown," "Unclimbed," and "Climbed." Additionally, a data entry error is corrected for Sharpu II (identified by peakid == "SPH2"), where the year of the first ascent is changed to 2018 to fix an obvious mistake.
-
-The second dataset cleaned is the expeditions dataset, which tracks various climbing expeditions.
-Here, the dataset is joined with the peaks dataset to bring in the peakname for each expedition using the peakid.
-Only the relevant columns are selected and renamed for clarity.
-The highpointmetres column is adjusted by setting any value of 0 to NA, as it likely indicates missing data.
-Categorical variables like terminationreason and season are recoded to make them more readable.
-The terminationreason column is recoded from numeric values (such as 0, 1, and 2) into descriptive labels like "Success (main peak)," "Bad weather," and "Accident." Similarly, the season column is recoded from numbers (0-4) into readable season names like "Spring," "Summer," "Autumn," and "Winter."
-
-The third dataset, members, contains information about individual climbers.
-This dataset is also joined with the peaks dataset to associate each climber's expedition with the corresponding peak name.
-As with the other datasets, only the relevant columns are retained and renamed.
-Any value of 0 in the age column is treated as missing data and changed to NA.
-The deathcause and injurytype columns, which are initially represented by numeric values (0-12), are recoded into readable categories like "AMS" (Acute Mountain Sickness) and "Exhaustion." The cleaning also involves handling information related to deaths and injuries.
-For climbers who have died, the deathcause and deathheightmetres are updated, and similarly, the injurytype and injuryheightmetres are updated for injured climbers.
 
@@
 
+## Future directions
+
+While our analysis provides insights into expedition outcomes based on team composition and basic environmental factors, a more comprehensive understanding of mountaineering risk and success requires integration of multiple observational data streams and advanced modeling techniques.
+
+Current environmental monitoring systems on major peaks like Everest provide crucial but limited data.
+Future research could benefit from a more extensive network of high-altitude weather stations providing real-time data on wind patterns, temperature gradients, precipitation, barometric pressure changes, and snow accumulation rates.
+Integration of satellite imagery could enhance these ground-based observations by tracking snow cover extent and depth, glacial movement patterns, ice fall dynamics, rock fall zones, and avalanche paths and debris patterns.
+
+Building on existing meteorological models, future research could develop route-specific risk models that integrate topographical data with weather patterns, modeling of snow loading and avalanche risk, ice fall prediction based on temperature patterns, and rock fall risk assessment incorporating freeze-thaw cycles.
+Physiological response simulations could model altitude acclimatization based on ascent profiles, energy expenditure predictions for different route sections, oxygen demand at various altitudes and exertion levels, and cold exposure risk assessment.
+
+Modern computational techniques could help synthesize multiple data streams, including historical expedition data, real-time weather observations, satellite imagery, individual climber biometrics, and equipment performance metrics.
+Machine learning models could then identify optimal weather windows for summit attempts, predict route-specific risks based on current conditions, suggest optimal acclimatization schedules, and recommend equipment configurations for specific conditions.
+
+This research would build upon existing studies like \[\@huey2020death\], who analyzed death rates on Everest, by incorporating their findings about team dynamics and oxygen use into more comprehensive risk models.
+The American Alpine Club's accident analysis \[\@aac2023accidents\] provides valuable historical data on accident patterns that could inform predictive models.
+
+Any proposed models would require extensive validation through historical data analysis, controlled testing in alpine environments, feedback from experienced mountaineers, and real-world implementation and monitoring.
+This research could lead to real-time risk assessment tools for expedition leaders, improved route planning and logistics, and better-informed decision-making about summit attempt timing, route selection, resource allocation, team composition, enhanced safety protocols, and more effective emergency response planning.
+
+Several challenges need to be addressed in this research, including data collection in extreme environments, integration of diverse data types, model validation in complex mountain environments, balancing predictive power with practical usability, cost and maintenance of monitoring systems, privacy considerations for individual climber data, and ethical implications of risk assessment tools.
+
+Future research should prioritize developing robust data collection systems and creating standardized risk assessment protocols while building user-friendly interfaces for practical application.
+Establishing international collaboration networks, integrating indigenous knowledge and local expertise, addressing ethical considerations in risk management, and studying the impact of climate change on mountaineering risk are also crucial areas for development.
+
+This comprehensive approach to observational data would significantly enhance our understanding of mountaineering risk factors and potentially lead to improved safety outcomes while maintaining the fundamental challenge and appeal of high-altitude climbing.
+The integration of modern technology with traditional mountaineering wisdom could create a more nuanced and effective approach to risk management in high-altitude expeditions.
+
 # References

### 2024-12-15 paper/paper.qmd
update

@@
 #| label: fig-year
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| fig-cap: "Distribution of years of expeditions"
 
@@
   ) +
-  theme_minimal()
+  theme_minimal() +
+   labs(
+    title = "Relationship between Hired staff and Success",
+    x = "Hired Staff",
+    y = "Outcome"
+  )
 
@@
 #| label: fig-ssdistribution
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| fig-cap: "Distribution of average age in expedition"
 ggplot(expeditions, aes(average_age)) +
@@
   ) +
-  theme_minimal()
-
-```
-
-\newpage
-
-```{r}
-#| warning: false
-#| message: false
-#| echo: false
-#| label: fig-age
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
-success_by_age <- expeditions |>
-  group_by(age_range) |>
-  summarize(success_rate = mean(success, na.rm = TRUE))
-ggplot(success_by_age,aes( x = age_range, y = success_rate))+
-  geom_bar(stat = 'identity')
+  theme_minimal() +
+   labs(
+    title = "Distribution of average age",
+    x = "Hired Staff",
+    y = "Outcome"
+  )
 
@@
 #| label: fig-age-success
-#| fig-cap: "Distribution of Himialayan peaks, mountains recorded are centered around 6200 M above sea level.."
+#| fig-cap: "Distribution of age and success has a nonlinear relationship, peaking at 35-40"
 
@@
   labs(
-    title = "Relationship between Height and Success",
+    title = "Relationship between age and Success",
     x = "Height",
@@
   labs(
-    title = "Histogram of Average Age in Expeditions",
-    x = "Average Age",
+    title = "Distribution of male proportion in Expeditions",
+    x = "Male proportion in expedition",
     y = "Count"
@@
   theme_minimal() +
-  labs(
-    title = "Distribution of male proportion in expeditions",
-    x = "Height",
-    y = "Outcome"
-  ) +
   theme(
@@
 
-success_rate <- expeditions |>
-  group_by(height_attempted)|>
-  summarize(success_rate = mean(success))
-ggplot(success_rate, aes(height_attempted,success_rate)) +
-  geom_point(alpha = 0.6, color = "#2c3e50") +
-  # Add trend line with confidence interval
-  geom_smooth(method = "loess", color = "#e74c3c") +
-  scale_y_continuous(limits = c(0, 1)) +
-  scale_x_continuous(limits = c(5500, 9000)) +
-  # Customize theme and labels
-  theme_minimal() +
-  labs(
-    title = "Relationship between Height of mountain and Success Rate",
-    x = "Sex Ratio",
-    y = "Success Rate",
-    caption = "Data source: Your dataset"
-  ) +
-  # Enhance readability
-  theme(
-    plot.title = element_text(size = 14, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10)
-  )
-####
-
 ggplot(expeditions, aes(x = height_attempted, y = as.numeric(success))) +
@@
   labs(
-    title = "Height of mountains attempted",
-    x = "Height in meters",
-    y = "Expeditions"
+    title = "Distribution of attempts on peaks",
+    x = "Attempts",
+    y = "Amount"
   ) +
@@
 #| echo: false
-#| fig-cap: "Amount of unclimbed Himalayan peaks"
+#| fig-cap: "Relationship of climb attempts on peak and success shows experience on a peak increases success rates."
 
@@
   labs(
-    title = "Relationship between Height and Success",
-    x = "Height",
+    title = "Relationship between previous attempts on the mountain and Success",
+    x = "Attempts",
     y = "Outcome"
@@
 ```
-
-This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
-Why do so many climbs fail however?
-The following graph shows the most common reasons.
 
@@
 A Binary Logistic Regression is a statistical method to predict binary outcomes like the success or failure of a climb based on both categorical and continuous variables.
-The Model uses nine independent demographic variables and the dependent variable is "success" The logistic regression model I will be using is: $$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{year} + \beta_7 \times \text{previousattempts}  + \beta_8 \times \text{sexratio} + \beta_9 \times \text{hiredstaff}$$
+The Model uses nine independent demographic variables and the dependent variable is "success" The intial logistic regression model I will be using is: $$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{year} + \beta_7 \times \text{previousattempts}  + \beta_8 \times \text{sexratio} + \beta_9 \times \text{hiredstaff}$$
+
+After removing variables through the process below we will end up with:
+
+$$\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{averageage} + \beta_2 \times \text{members} + \beta_3 \times \text{oxygenused} + \beta_4 \times \text{season} +  \beta_5 \times \text{heightattempted} + \beta_6 \times \text{previousattempts}$$
 
@@
 
-\newpage
-
-This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
+@tbl-summary1 shows the influence (top number) and P-Value (bottom number in brackets) of each variable.
 For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship.
@@
 Because we have so many statistically insignificant variables at a 5% uncertainty we should try to remove them to see the impact it has on the model
+
+\newpage
 
@@
 Therefore, the simpler (reduced) model is likely preferable because it achieves similar performance with fewer parameters.
-\newpage \## Results {#sec-results}
+\newpage
+
+# Results {#sec-results}
 

### 2024-12-15 paper/paper.qmd
updating bibliography

@@
 
-The datasets were cleaned and analysed using the statistical programming software `R` [@citeR] along with the help of `tidyverse` [@citeTidyverse], `knitr` [@citeKnitr], `ggplot2` [@citeGgplot], `here` [@citeHere], `dplyr` [@citeDplyr], `rstanarm` [@citeRstanarm], `arrow` [@citeArrow] and `kableExtra` [@citeKableExtra].
+The datasets were cleaned and analysed using the statistical programming software `R` [@citeR] along with the help of `tidyverse` [@citeTidyverse], `knitr` [@citeKnitr], `ggplot2` [@citeGgplot], `here` [@citeHere], `dplyr` [@citeDplyr], `rstanarm` [@citeRstanarm],`stringr` [@stringr], `brms` [@brms], `rsample` [@rsample], `stats` [@stats], `MASS` [@MASS], and `pROC` [@pROC], `arrow` [@citeArrow] ndefinedand `kableExtra` [@citeKableExtra].
 

### 2024-12-15 paper/paper.qmd
Spell check

@@
 This paper uses data sourced from @Cookson, based on The Himalayan Database [@Salisbury] about expeditions from 1905-2019 to analyze which factors help or deter climbers from succeeding.
-We use a binomial logistic regression model to determine the probability of success based on a variety of demographic, enviromental, and organizational factors.
+We use a binomial logistic regression model to determine the probability of success based on a variety of demographic, environmental, and organizational factors.
 
@@
 Importantly it doesn't capture the same individual in different expeditions instead assigning two `member_ids`.
-It connects to Expeditions with `expeditions_id` and contains alot of information about the specific member during the expedition.
+It connects to Expeditions with `expeditions_id` and contains a lot of information about the specific member during the expedition.
 Most importantly for this data we use average sex and age of all the members in each expeditions for the model.
@@
 #| fig-height: 5
-#| fig-cap: "A slight majority of Expeditions succeed, with the most common reasons expeditions fail being enviromental issues"
+#| fig-cap: "A slight majority of Expeditions succeed, with the most common reasons expeditions fail being environmental issues"
 
@@
 #| label: fig-members
-#| fig-cap: "Most expeditions have less than 3 members showing an exponetial decrease with an increase in members. The red lines indicate member amounts that have only had one expedition"
+#| fig-cap: "Most expeditions have less than 3 members showing an exponential decrease with an increase in members. The red lines indicate member amounts that have only had one expedition"
 
@@
 #| label: fig-sex-success
-#| fig-cap: "Relationship of male proportion of expedition and succes is a small positive relationship possibly explained by other factors"
+#| fig-cap: "Relationship of male proportion of expedition and success is a small positive relationship possibly explained by other factors"
 
@@
 This is when you remove variables from the model one at a time until it not longer improves the AIC.
-We see the RMSE doesnt change, meaning it retains the same accuracy as the full model and the AIC is decreased.
+We see the RMSE doesn't change, meaning it retains the same accuracy as the full model and the AIC is decreased.
 
